Opposing chemosensory functions of closely related gustatory receptors

Version 1

Abstract

Abstract

Most animals possess functionally distinct population of taste cells, expressing receptors that are tuned to compounds of different valence. This organizational feature allows for discrimination between chemicals associated with different taste modalities and facilitates sensing of foods contaminated with toxic chemicals. In the fruit fly D. melanogaster, primary sensory neurons express taste receptors that are tuned to distinct chemicals, thereby activating neural ensembles that elicit either feeding or avoidance behavior. Members of a family of ligand gated receptor channels, the Gustatory receptors (Gr), play a central role in these behaviors. In general, closely related, evolutionarily conserved Gr proteins are co-expressed in the same type of taste neurons, tuned to chemically related compounds and therefore triggering the same behavioral response. Here, we report that members of the Gr28 subfamily are expressed in largely non-overlapping sets of taste neurons in Drosophila larvae, detect chemicals of different valence and trigger opposing feeding behaviors. We determined the intrinsic properties of Gr28 neurons by expressing the mammalian Vanilloid Receptor (VR1), which is activated by capsaicin, a chemical to which wild type Drosophila larvae do not respond. When VR1 is expressed in Gr28a neurons, larvae become attracted to capsaicin, whereas expression of VR1 in Gr28bc neurons triggers avoidance to capsaicin. Thus, the Gr28 proteins, which have been evolutionarily conserved over the last 65 million years in insects, represent the first taste receptor subfamily in which specific members mediate behavior with opposite valence. We also identified denatonium benzoate, a synthetic bitter compound that shares structural similarities with natural bitter chemicals, as a ligand for a receptor complex containing a Gr28bc or Gr28ba receptor subunit. Alphafold structure prediction, combined with the limited sequence conservation in the putative binding pockets of various Gr28 proteins, creates a theoretical framework for targeted in vivo structure function studies to precisely map residues critical for ligand recognition.

Introduction

Meaningful animal behavior is established through cooperative engagement of multiple sensory systems. In many insects, the chemosensory system plays a central role in such integration processes. The fruit fly Drosophila melanogaster has served as the primary insect model system for elucidating the molecular basis and neural circuitry of both olfaction and taste, by virtue of the vast genetic resources, amenability to both neurophysiological recording and live imaging, and simple yet powerful behavioral assays, allowing investigators to link genes to chemosensory behavior and neural activity (Montell, 2021).

The gustatory system in Drosophila is characterized by several insect specific features with regard to anatomy and the molecular nature of taste receptors. For example, taste cells are primary sensory neurons, with dendritic projections that express taste receptors, while long axons project that convey taste information directly to the brain. In adult flies, these neurons, referred to as Gustatory Receptor Neurons (GRNs) are distributed across several appendages, such as labial palps, legs and presumably the antennae. Additionally, some GRNs reside internally, arranged in cell clusters along the pharynx (Joseph and Carlson, 2015)(Amrein, 2016)(Scott, 2018). Similarly, Drosophila larvae have numerous structures located both on the head surface as well as internally along the larval pharynx, that assess chemical compounds prior and during ingestion, respectively (Kwon et al., 2011)(Apostolopoulou et al., 2015)(Rist and Thum, 2017). Like most insects, Drosophila employs taste receptors encoded by two major gene families, the Gustatory receptor (Gr) and the Ionotropic Receptor (IR) genes, to sense all chemicals, such as appetitive food compounds, noxious and toxic chemicals as well as pheromones. Both Gr and IR based receptors are thought of form complexes composed of several different subunits, partial composition of which is known for only a few Gr-based taste receptors for sugars and bitt er compounds (Kwon et al., 2007)(Jiao et al., 2008)(Fujii et al., 2015)(Shim et al., 2015)(Gonzales et al., unpublished), as well as gaseous carbon dioxide (Jones et al., 2007) and even fewer IR-based fatty acid and carboxylic acid receptors (Ahn et al., 2017)(Tauber et al., 2017)(Chen and Amrein, 2017). Interestingly, at least some GRNs express members of both families, although no evidence for receptor complexes consisting of both IR and Gr subunits has been reported to date.

The Gr genes represent the largest taste receptor gene family in insects. In D. melanogaster, it comprises 60 genes predicted to encode 68 proteins, expression of which has been characterized (Kwon et al., 2011)(Weiss et al., 2011)(Ling et al., 2014)(Choi et al., 2016)(Rist and Thum, 2017). Several Gr genes have been functionally characterized in adult flies using genetic mutations, combined with either electrophysiological recordings, Ca2+ imaging studies and behavioral analyses (reviewed by Montell, 2021) but only a few have been studied in larvae (Mishra et al., 2013)(Apostolopoulou et al., 2016)(Choi et al., 2016)(Mishra et al., 2018)(Choi et al., 2020). The Gr28 gene subfamily is of particular interest for a number of reasons: First, it is one of the most conserved Gr subfamily, homologs of which can be found across all insect families and even more distant arthropods (Eyun et al., 2017)(Suzuki et al., 2018)(Fujii et al., 2023). Second, the Gr28 genes are expressed not only in the gustatory system, but in many other organs, especially the central nervous system (CNS) and non-chemosensory neurons of the peripheral nervous system (PNS), suggesting that they have functions beyond gustation and are important to sense chemical signals unrelated to food (Thorne and Amrein, 2008). Notably, evidence for such roles has been reported before any direct link to gustatory perception was discovered. Ni and colleagues showed that Gr28bd is essential for high temperature avoidance in flies (Ni et al., 2013), while Xiang and collaborators found that Gr28 mutant larvae were deficient in UV light avoidance (Xiang et al., 2010). Third, the only gustatory function for any Gr28 gene in larvae thus far is sensing of ribonucleosides and RNA in larvae (Mishra et al., 2018), which was recently established as a novel appetitive taste quality common to a broad range of dipteran insects (Fujii et al., 2023).

Here, we present a functional characterization of the Gr28 genes and the respective GRNs in Drosophila larvae. Four of the six Gr28 genes (Gr28a, Gr28ba, Gr28bc and Gr28be) are expressed in the larval taste system, subdivided into two, functionally distinct neuronal ensembles. The first such ensemble is defined by expression of Gr28bc and comprises four neurons, two of which also express Gr28ba, Gr28bc, Gr28be and the bitter core receptor gene Gr66a, while the other two neurons express only Gr28bc, Gr28a and Gr66a. The second ensemble is characterized by 12 GRNs that express Gr28a, including the pair of GRNs expressing the Gr28b genes. When the mammalian Vanilloid Receptor 1 (VR1) was expressed under the control of specific GAL4 drivers, we found that Gr28a and Gr28bc neurons mediate opposing taste behavior in the presence of capsaicin, the ligand for VR1. Specifically, Gr28a:VR1 larvae show strong attraction for capsaicin, while Gr28bc:VR1 larvae show strong avoidance of capsaicin. Furthermore, inactivation experiments reveal that the Gr28bc GRNs are necessary to sense bitter compounds, specifically denatonium, quinine, lobeline, and caffeine. However, Gr28 mutant larvae fail to avoid only denatonium, but none of the other bitter compounds, a phenotype rescued by expression of Gr28bc or Gr28ba. Thus, to our knowledge, the related Gr28 genes represent the first subfamily where members mediate functionally opposing taste behaviors.

Results

Expression of the Gr28 genes in the larval taste organs

The peripheral chemosensory system of the larvae is subdivided into bilaterally symmetrical, ‘external’ and ‘internal’ taste organs (Stocker, 2008). The three external organs reside near the tip of the head and organized as paired ganglia, the Dorsal, Terminal and Ventral Organ Ganglia (DOGs, TOGs and VOGs) that house the GRN cell bodies, each neuron projecting a dendrite to the sensory organ structures (DO, TO and VO) at the head surface, while carrying information via their axons to the subesophageal zone (SEZ) in the brain (Figure 1A). The DOG harbors 21 olfactory neurons and 18 presumptive taste neurons. For clarity, numbers below refer to neurons on one side of the bilaterally symmetrical structures of all organs, thus the total neuron number in a larva is double that number. The presumptive GRNs located in the DOG fall into two distinct groups, based on their dendritic projections: 11 presumptive GRNs, four of which were shown to express Gr genes (Apostolopoulou et al., 2015), extend dendritic projections to the base of the DO, while 7 neurons project dendrites to the TO (the dorsolateral group, Figure 1A)(Kwon et al., 2011). The TOG contains 30 neurons, all sending dendrites to the TO (the distal group). The internal taste structures, referred to as the dorsal/ventral pharyngeal sense organ (DPS/VPS, 33 neurons), and the posterior pharyngeal sense organs (PPS, 6 neurons) are located along the pharynx and sense chemicals as they pass towards the digestive system. We note that not all these neurons are confirmed GRNs, either by function or expression of Gr or IR genes, albeit based on their location and anatomy, most are thought to be GRNs (Stewart et al., 2015)(Rist and Thum, 2017)(Sánchez-Alcañiz et al., 2018).

Expression of the Gr28 genes in the larval sensory organs

A) Schematic representation of larval chemosensory system (Kwon et al., 2011). Three external sensory organs (Dorsal Organ, DO; Terminal Organ, TO; Ventral Organ, VO) hold collectively the dendritic projections of neuronal cell bodies in the respective ganglia (DOG, TOG and VOG). Three clusters of sensory neurons (Dorsa/Ventral Pharyngeal Sensory organ, D/VPS; and Posterior Pharyngeal Sensory organ, PPS) align the pharynx. The antennal nerve (AN) connects the DO neurons to the subesophageal zone (SEZ). The TO and VO neurons project along the maxillary nerve (MN) to the SEZ. The DPS and PPS neurons innervate to the SEZ via labral nerve (LRN). The VPS neurons project to the SEZ through the labial nerve (LBN). Note that olfactory neurons are omitted in the schematic here and in B (below).

B) Expression of the Gr28 genes in the external sensory organs. Note that images show only one of the bilaterally symmetrical organs. Possible co-expression between Gr66a and different Gr28b genes (in B and C; left panels) was assessed using GAL4 and LexA drivers for Gr28 genes and Gr66a, respectively. Expression summary (diagram on the right) shows only one side of the bilaterally symmetrical organs. The Gr28b genes and Gr28a are expressed in different neurons. Likewise, no co-expression is observed between Gr66a (marker for bitter GRNs) and Gr28a, whereas the Gr28b genes are co-expressed in a subset of Gr66a neurons.

C) Expression of the Gr28 genes in internal taste organs. Note that images for the DPS/VPS show both side of the bilaterally symmetrical ganglion, while images of the PPS show only one side of the bilaterally symmetrical organ. Gr28a and Gr66a are partially co-expressed, but each gene is also expressed exclusively in a subset of neurons in the PPS. Gr28bc but none of the other Gr28b genes is co-expressed with Gr66a in a single neuron in the DPS/VPS. Expression summary (diagram on the right) shows only one of the bilaterally symmetrical organs. Immunostaining with anti-GFP (green) and anti-mCD8 (magenta) antibodies was carried out on whole-mount preparations from larvae heads of the following genotypes: (1) UAS-mCD8:RFP lexAop-rCD2:GFP;Gr66a-LexA/Gr28a-GAL4 or Gr28bc-GAL4 or Gr28be-GAL4 (2) UAS-mCD8:RFP lexAop-rCD2:GFP;Gr66a-LexA/+; Gr28ba-GAL4/+. Asterisks refer to a GRN expressing both Gr66a and the indicated Gr28 gene. Scale bars are 5 μm.

Our initial expression analysis revealed that four of the six Gr28 genes (Gr28a, Gr28ba, Gr28bc and Gr28be) are expressed in larval taste organs, in addition to cells in the gut, the brain and non-chemosensory cells of the larvae (Mishra et al., 2018). To further delineate the putative chemosensory roles of these genes, we performed a more detailed co-expression analysis between Gr28 genes and with respect to the bitter receptor gene Gr66a, using the bimodal expression systems GAL4/UAS to label Gr28 expressing neurons and LexA/lexAop to mark Gr66a neurons (Figure 1B and 1C; note that the neuron numbers below refer to the number present in one of each paired ganglion). In the TOG, we found expression of all four Gr28 genes, along with that Gr66a, which was expressed in 3 or 4 neurons (one neuron exhibits often stronger expression of Gr66a than the two or three remaining neurons). This number slightly smaller than the 6 neurons reported using a distinct Gr66a-GAL4 driver (Kwon et al., 2011). Gr28ba, Gr28bc and Gr28be are co-expressed in a single neuron with Gr66a (Figure 1B). In contrast, Gr28a is expressed in a distinct TOG neuron than Gr66a (Figure 1B), and by extension, in a different neuron than the co-expressed Gr28b genes, which we independently confirmed using a Gr28bc-QF driver (Figure 1-figure supplement 1). In the DOG, we find only a single Gr28a neuron, while none of the Gr28b genes, or Gr66a, is expressed there (Figure 1B). Expression in the internal sensory organs was similarly complex. Gr66a was found in one neuron in the DPS/VPS, co-expressed with Gr28bc as well as Gr28a (Figure 1C). None of the Gr28b genes is expressed in the PPS, but two neurons express Gr28a, one of which also expresses Gr66a (Figure 1C). In summary, the internal sensory organs can be subdivided in three distinct groups: Gr66a/Gr28b/Gr28a positive neurons, Gr66a/Gr28a positive neurons and Gr28aonly neurons, which contrasts expression in the external sensory organs where Gr28a and the Gr28b genes are mutually exclusive.

Subsets of Gr28 neurons mediate opposing feeding behaviors

We previously showed that at least one of the six Gr28 genes is necessary for feeding attraction to RNA, ribonucleosides and ribose using a well-established two choice feeding assay (Mishra et al., 2018). Specifically, larvae homozygous for a Gr28 null allele, a deletion of the entire Gr28a gene and most of the coding region of all Gr28b genes (ΔGr28), lose their ability to sense these compounds, a phenotype that is restored when single UAS-Gr28 genes are expressed in Gr28a neurons using Gr28a-GAL4. This observation and the largely non-overlapping expression of Gr28a and the Gr28b genes suggests that respective neurons represent functionally distinct entities. To investigate this possibility, we took advantage of the mammalian Vanilloid Receptor 1 (VR1), a TRP channel that is activated by capsaicin (Caterina et al., 1997). Drosophila have no VR1 like-gene in their genome and do not respond to capsaicin behaviorally, but flies are attracted to this chemical when a modified VR1 gene (VR1E600K) is expressed in sweet taste neurons (Marella et al., 2006). Thus, we expressed VR1E600K (henceforth referred to UAS-VR1) under the control of the four Gr28-GAL4 drivers (Gr28a-GAL4, Gr28ba-GAL4, Gr28bc-GAL4 and Gr28be-GAL4) in larvae and tested their response to capsaicin using the two-choice preference assay. Additionally, we expressed UAS-VR1 in bitter neurons as well as sweet neurons using respective GAL4 drivers (Gr66a-GAL4 and Gr43aGAL4) (Scott et al., 2001) (Miyamoto et al., 2012). Consistent with the observation in adults (Marella et al., 2006), w1118 control larvae and larvae carrying UAS-VR1 or Gr-GAL4 genes alone were unresponsive to 0.1 mM or 0.5 mM capsaicin (Figure 2A). However, larvae expressing UAS-VR1 under the control of Gr28a-GAL4 or Gr43a-GAL4 showed strong appetitive responses to both capsaicin concentrations (Figure 2A). In contrast, larvae expressing UAS-VR1 under the control of Gr28bc-GAL4 or Gr66a-GAL4 showed robust avoidance behavior to both concentrations of capsaicin. Lastly, expression of UAS-VR1 with the Gr28ba-GAL4 or Gr28be-GAL4 driver, which is co-expressed in a single TO neuron with Gr28bc (Figure 1B), did cause neither attraction to nor avoidance of capsaicin.

Taste behavior of larvae mediated by different Gr28 GRNs

A) Gr28a GRNs mediate capsaicin preference, while Gr28bc GRNs elicit capsaicin avoidance in larvae. Control larvae (w1118, UAS-VR1/+ and Gr (indicated Gr genes)-GAL4/+) show no preference for or avoidance to 0.1 mM (top) or 0.5 mM (bottom) capsaicin. When expressing UAS-VR1 in Gr28a GRNs (Gr28a-GAL4/UAS-VR1), larvae display robust taste preference for capsaicin for both 0.1 and 0.5 mM capsaicin. In contrast, when expressed in Gr28bc GRNs (Gr28bc-GAL4/UAS-VR1), larvae strongly avoid both 0.1 and 0.5 mM capsaicin. Neither avoidance nor preference was observed when VR1 was expressed in the Gr28ba or Gr28be GRNs. For comparison, expression of VR1 in sweet taste neurons (Gr43aGAL4/UAS-VR1) or bitter taste neurons (Gr66a-GAL4/UAS-VR1) led to strong preference for or strong avoidance of capsaicin. Each bar represents the mean ± SEM of P.I. (n = 11-42 assays). The taste behavior of Gr-GAL4 > UAS-VR1E600K larvae is compared with 3 controls (w1118, UAS-VR1E600K/+ & Gr-GAL4/+) by using one-way ANOVA with Bonferroni’s multiple comparison tests (p < 0.05), whereby different letters indicate statistically significant difference. Dashed lines delineate groups for ANOVA. A short solid line indicates control groups.

B) Gr28aonly GRNs are necessary and sufficient for inducing capsaicin preference of larvae. When VR1 expression is restricted to GRNs not co-expressing Gr66a or Gr28bc (Gr28aonly GRNs) by means of lexAop-GAL80 under control of Gr66a-LexA, larvae show strong preference for capsaicin. Each bar represents the mean ± SEM of P.I. (n = 13-18 assays). Bars with different letters are significantly different (one-way ANOVA with Bonferroni’s multiple comparison tests (p < 0.05). Fly genotypes: wild-type: w1118 (black), w1118;UAS-VR1E600k/+ (light gray), w1118;Gr28a-GAL4/+ (dark grey), w1118;Gr28a-GAL4/UAS-VR1E600K (white), w1118;+;Gr28ba-GAL4/+ (dark grey), w1118;UAS-VR1E600K/+;Gr28ba-GAL4/+ (white), w1118;Gr28bc-GAL4/+ (dark grey), w1118;Gr28bc-GAL4/UAS-VR1E600K (white), w1118;Gr28be-GAL4/+ (dark grey), w1118;Gr28be-GAL4/UAS-VR1E600K (white), w1118;Gr43aGAL4/+ (dark grey), w1118;Gr43aGAL4/UAS-VR1E600K (white), w1118;Gr66a-GAL4/+ (dark grey), w1118;Gr66a-GAL4/UAS-VR1E600K (white), w1118;UAS-VR1E600k/+;lexAop-GAL80/+ (light gray), w1118;Gr66a-LexA/+;Gr28a-GAL4/+ (dark grey) and w1118;Gr66a-LexA/UAS-VR1E600K;Gr28a-GAL4/lexAop-GAL80 (white).

Expression difference between Gr28bc and Gr28ba/be is a single neuron in the DPS/VPS (Figure 1C), a neuron required to mediate avoidance of capsaicin in Gr28bc-GAL4, UAS-VR1 larvae. Even though this GRN also expresses Gr28a, it appears unlikely that it is necessary to mediate appetitive behavior as well. Indeed, we could further restrict the necessary set of appetitive neurons to four Gr28a positive neurons (Gr28aonlyGRNs), as suppression of VR1 expression in Gr66a GRNs (one being the Gr28bc expressing GRN in the DPS/VPS) did not affect capsaicin preference (Figure 2B). We further note that activation of the two Gr28bc neurons by capsaicin drives repulsion to the same degree as when all seven Gr66a expressing neurons are activated. Taken together, this analysis indicates that Gr28a-GAL4 is expressed in appetitive inducing neurons, while Gr28bc-GAL4 neurons mediate avoidance behavior.

Wild type larvae, like adult flies, avoid bitter compounds (Apostolopoulou et al., 2015) (Apostolopoulou et al., 2016) (Choi et al., 2016) (van Giesen et al., 2016) (Choi et al., 2020). If Gr28bc neurons are indeed central to the avoidance of these chemicals in larvae, we expect that eliminating their function would result in loss of larval avoidance behavior to bitter chemicals. Thus, we inactivated neuronal subsets using the potassium inward rectifying channel Kir2.1 (Baines et al., 2001) (Paradis et al., 2001). In contrast to larvae with an intact taste neuron repertoire, larvae in which two Gr28bc neurons were inactivated no longer avoided any of these bitter compounds (Figure 3). A more subtle phenotype was observed when the single Gr28be neurons were inactivated, eliminating lobeline and caffeine avoidance, but leaving denatonium and quinine repulsion intact, a finding consistent with previous studies (Choi et al., 2020). These data suggest that two pairs of neurons, one in the TO and one in the D/VPS are necessary and sufficient for avoidance of all four bitter tasting chemicals.

The Gr28b neurons mediate avoidance behavior to bitter compounds

Inactivation of Gr28bc GRNs (Gr28bc-GAL4/UAS-Kir2.1) elicits significantly reduced avoidance of larvae to all four bitter compounds tested: denatonium, quinine, lobeline and caffeine, while control larvae carrying either the driver or the reporter only showed strong avoidance of these compounds. In contrast, flies with inactivated Gr28be GRNs (Gr28be-GAL4/UAS-Kir2.1) still avoid denatonium and quinine (top), but no longer avoid lobeline and caffeine (bottom). Each bar represents the mean ± SEM of P.I. (n = 11-20 assays). The taste behavior of Gr-GAL4 > UAS-Kir2.1 larvae is compared with 2 controls (UAS-Kir2.1/+ & Gr28b-GAL4/+) by using Kruskal-Wallis test by ranks with Dunn’s multiple comparison tests (p < 0.05). Bars with different letters are significantly different. Dashed lines delineate groups for ANOVA. A short solid line indicates a control. Fly genotypes: w1118; UAS-Kir2.1/+ (light grey), w1118; Gr28bc-GAL4/+ (dark grey), w1118; Gr28bc-GAL4/UAS-Kir2.1 (white), w1118; Gr28be-GAL4/+ (dark grey), w1118; Gr28be-GAL4/UAS-Kir2.1 (white).

Gr28bc and Gr28ba are subunits of a taste receptor complex for denatonium

We examined whether any of the Gr28 proteins is part of a taste receptor complex detecting these bitter chemicals (Figure 4). Surprisingly, only avoidance of denatonium was affected in larvae without any functional Gr28 gene (ΔGr28), while responses to quinine, lobeline or caffeine was not significantly affected. Moreover, avoidance to caffeine increased significantly (Figure 4A). We next tested which Gr28 protein is necessary for denatonium avoidance and expressed individual Gr28 genes under the control of the Gr28bc-GAL4 driver. Only when either Gr28ba or Gr28bc was expressed in these neurons, denatonium avoidance was fully recovered, whereas expression of Gr28bb, Gr28bd, Gr28be or Gr28a failed to do so (Figure 4B). This observation indicates that despite the high level of similarity between these receptors, recognition of denatonium is dependent on specific Gr28b subunits, features that are present in Gr28ba and Gr28bc, but not any of the other Gr28 proteins.

Role of individual Gr28 genes in bitter taste avoidance

A) Some of the Gr28 genes are strictly required for sensing denatonium. Wild-type (w1118) larvae, but not Gr28 mutant larvae, strongly avoid denatonium. In contrast, Gr28 mutant (w1118;ΔGr28/ΔGr28) larvae do not show significantly reduced avoidance of quinine, lobeline and caffeine. Each bar represents the mean ± SEM of P.I. (n = 12-22 assays). Asterisks indicate a significant difference between the Gr28 mutant and wild-type larvae (Two-tailed, Mann-Whitney U test, **** p < 0.0001, *** p < 0.001, ns: not significant).

B) Single Gr28b genes can rescue avoidance response to denatonium when expressed in Gr28bc neurons of Gr28 mutant larvae. Each bar represents the mean ± SEM of P.I. (n = 11-22 assay). The behavior of Gr28 mutant larvae expressing UAS-Gr28 transgenes under control of the Gr28bc-GAL4 driver was compared to Gr28+/+control (w1118, filled bar), and three Gr28 mutant controls (ΔGr28, ΔGr28 plus driver, and ΔGr28 plus respective UAS-Gr28 reporter, plain bar) using Kruskal-Wallis test by ranks with Dunn’s multiple comparison tests (p < 0.05). Bars with different letters are significantly different. Fly genotypes: wild-type: w1118(black), mutants: w1118;ΔGr28/ΔGr28, w1118;ΔGr28/ΔGr28 Gr28bc-GAL4, w1118;ΔGr28/ΔGr28;UAS-Gr28 (indicated Gr28 genes)/+ and w1118;ΔGr28/ΔGr28 UAS-GCaMP6m;UAS-Gr28 (for Gr28bb or Gr28bc genes)/+ (white), rescues: w1118;ΔGr28/ΔGr28 Gr28bc-GAL4;UAS-Gr28 (indicated Gr28 genes)/+ and w1118;ΔGr28 UAS-GCaMP6m/ΔGr28 Gr28bc-GAL4;UAS-Gr28 (for Gr28bb or Gr28bc genes)/+ (grey).

Previous work had established that Gr66a is required for caffeine avoidance, in both larvae and adult flies (Moon et al., 2006)(Lee et al., 2009)(Apostolopoulou et al., 2016). We confirmed this, as Gr66a mutant larvae show little if any avoidance of caffeine, while avoidance of lobeline was not affected (Figure 4-figure supplement 1). Somewhat surprisingly, avoidance of denatonium and quinine increased significantly. This is likely a result of the heteromeric composition of bitter taste receptor complexes, whereby absence of Gr66a allows Grs that tuned to denatonium (such as Gr28bc or Gr28ba) and quinine be incorporated into multimeric receptor complexes.

Finally, we investigated neuronal responses in larvae expressing the Ca2+ indicator GCaMP6m in Gr28bc-GAL4 GRNs (Figure 5). We developed a whole larvae imaging preparation, whereby larvae were placed in an “imaging chamber” to minimize head movements (Figure 5A), and visualized neural activity in real time (Chen et al., 2013) upon exposure to the four bitter compounds, as well as sucrose, ribose and fructose (Figure 5B to 5D). All bitter compounds elicited rapid Ca2+ increases in Gr28bc GRNs in the TO, while none of the sugars did (Figure 5C and 5D). When neural activity was recorded in ΔGr28 homozygous mutant larvae (Figure 5E), cellular responses to denatonium and quinine were severely reduced, while responses to both caffeine and lobeline were not affected. Consistent with our behavioral analysis (Figure 4B), re-expression of either Gr28ba or Gr28bc, but not Gr28bb, Gr28bd, Gr28be or Gr28a rescued cellular response to denatonium, but not to quinine (Figure 5E). Together, these experiments identified Gr28bc and Gr28ba as redundant subunit of a denatonium receptor complex, a complex that does not require Gr66a or any of the other Gr28b subunits.

Cellular activation of Gr28bc GRNs by bitter compounds that require Gr28ba or Gr28bc function.

A)Diagram of Ca2+ imaging experimental set up.

B) Representative still images of Ca2+ response in the Gr28bc expressing TO neuron. Ca2+ responses of the Gr28bc GRNs upon stimulation with indicated ligands. ΔF indicates the changes in fluorescence light intensity of the cell body before/after ligand application.

C & D) Representative traces (C) and quantified Ca2+ responses (D) of the Gr28bc GRNs after stimulation with indicated ligands. Fly genotype: w1118; Gr28bc-GAL4/UAS-GCaMP6m. Each bar represents the mean ± SEM of Ca2+ imaging with 12 – 16 larvae. Asterisks indicate a significant difference between carrier (water) and indicated ligands (Two-tailed, Mann-Whitney U test, *** p < 0.001, ns: not significant).

(E) Neurons of larvae lacking the Gr28 genes exhibit significantly reduced responses to denatonium and quinine. The Gr28bc expressing TO GRN of Gr28 mutant larvae (ΔGr28) display significantly reduced Ca2+ responses to denatonium and quinine but not to lobeline or caffeine when compared to Gr28bc expressing TO of wild type controls. Responses to denatonium, but not quinine, is restored by expression of single Gr28b (Gr28ba or Gr28bc) genes (grey). Larvae genotypes: Gr28+ control (black bar): w1118; Gr28bc-GAL4/UAS-GCaMP6m. Gr28- control (white bar): w1118; ΔGr28 Gr28bc-GAL4/ΔGr28 UAS-GCaMP6m. Each bar represents the mean ± SEM with 13 – 16 larvae. Asterisks indicate a significant difference between Gr28+control and Gr28 mutant (ΔGr28) larvae (Two-tailed, Mann-Whitney U test, *** p < 0.001, ** p < 0.01 ns: not significant).

(F and G) Gr28bc or Gr28ba rescues denatonium responses in GRNs of ΔGr28 larvae. Expression of Gr28bc or Gr28ba in under control of Gr28bc-GAL4 restores responses to denatonium, but not quinine in GRNs of ΔGr28 larvae. Each bar represents the mean ± SEM of Ca2+ imaging with 12 – 17 larvae. The Ca2+ responses of Gr28 mutant larvae expressing UAS-Gr28 transgenes under Gr28bc-GAL4 driver is compared to Gr28+ (black) and Gr28-(white) controls by using Kruskal-Wallis test by ranks with Dunn’s multiple comparison tests (p < 0.05). Bars with different letters are significantly different. Dashed lines delineate groups for ANOVA. A short solid line indicates control groups. Fly genotypes: Gr28+ control (black bar): w1118; Gr28bc-GAL4/UAS-GCaMP6m. Gr28- control (white bar): w1118; ΔGr28 Gr28bc-GAL4/ΔGr28 UAS-GCaMP6m. Gr28 rescues (grey bar): w1118; ΔGr28 Gr28bc-GAL4/ΔGr28 UAS-GCaMP6m; UAS-Gr28/+. Concentrations of ligands were 100 mM for sugars, 50 mM for caffeine and 5 mM for denatonium, quinine and lobeline.

Discussion

The Gr28 receptors comprise one of the few Gr subfamilies conserved across diverse insect species (Engsontia and Satasook, 2021)(Agnihotri et al., 2016)(Yu et al., 2023). Yet, they were the least characterized when compared to other conserved subfamilies, such as the sugar receptors (Gr5a, Gr61a and Gr64a-f), the carbon dioxide receptors Gr21a and Gr63a or the bitter taste receptors. The only ligands associated with the Gr28 proteins were ribonucleosides and RNA, which are appetitive nutrients for larvae and are mediated by Gr28a neurons (Mishra et al., 2018). Indeed, RNA has been found to be an appetitive taste ligand across many dipteran insects, including mosquitoes, and we showed that Gr28 homologs of both A. aegypti and A. gambiae can rescue the preference for RNA and ribose when expressed in Gr28a neurons of ΔGr28 mutant larvae (Fujii et al., 2023). Here, we established a precedent for members of the same subfamily for mediating both appetitive and repulsive behavior and being expressed in largely non-overlapping sets of GRNs. In flies, members of other conserved Gr subfamilies are either precisely (Gr21a and Gr63a)(Jones et al., 2007)(Kwon et al., 2007), or largely (sugar Gr genes) (Fujii et al., 2015) co-expressed in one type of GRNs, forming multimeric receptors that mediate a single type of behavior.

Our expression analysis in larvae is consistent with earlier reports, despite some variation in number of neurons expressing Gr66a and Gr28b (Kwon et al., 2011)(Choi et al., 2016), which is likely due to the use of different GAL4 drivers and/or variability in expression. Importantly, all Gr28b genes are co-expressed with the bitter marker Gr66a (Figure1), while Gr28a is found in a largely, but not entirely distinct set of GRNs. Based on our VR1-capsaicin experiments, the GRNs sharing expression of both Gr28a and Gr66a (and by extension some Gr28b genes) are critical for avoidance, and not required for appetitive behavior (Figure 2).

Distinct functions of small set of GRNs expressing specific Gr28 subunits

A key finding presented of this study is the observation that Gr28aonly and Gr28bc neurons dictate distinct behavioral programs, the former representing an ensemble of neurons that instruct a larvae to “go towards” chemical source and consume it, while the latter do the opposite (Figure 2). The set of “go-away” GRNs co-expressing Gr28bc and Gr66a is remarkably small, consisting of only two pairs, one in the TO and the other in the DPS/VPS. It seems likely that this is the smallest, minimal subset of neurons sufficient for avoidance, as expression of V1R in only the TO pair (under the control of either Gr28ba-GAL4 or Gr28be-GAL4) has no behavioral effect when challenged with capsaicin. The “go-to” neurons are characterized by expression of Gr28a and represent a slightly larger set of four GRN pairs (Gr28aonly only GRNs). Thus, we propose that the minimal requirement to induce “go-to” and “go-away” behavior is defined by distinct sets of GRNs, and that each appears to be composed of neurons located in both external as well as the internal taste organs (Figure 6).

Role of different GRN subsets in taste behavior of larvae

GRNs mediating avoidance behavior express Gr28bc, while GRNs necessary for feeding preference express Gr28a alone (Gr28aonly GRNs). Note that each ensemble is composed of at least a pair of neurons located in the external taste organs and a pair of neurons in the internal taste organs.

Co-expression of a subunit for the bitter taste receptors Gr28bc/Gr28ba and Gr28a, a receptor for RNA, a nutrient essential for larval growth (Mishra et al., 2018) in bitter taste neurons raises interesting questions about the function of Gr28a. As these neurons are necessary for repulsion, we propose that Gr28a might serve as a subunit along with other Grs (possibly Gr28bc and/or Gr66a) in heteromeric bitter taste receptor complexes, detecting yet other bitter compounds. In “go-to”, Gr28aonly GRNs, Gr28a is likely to form a homomultimeric receptor, since its expression in fructose sensing GRNs confers RNA recognition to these neurons, both in cellular and behavioral assays (Mishra et al., 2018). We note that the sweet taste receptor Gr5a, a subunit of a multimeric receptor in sweet taste GRNs in the labial palps of flies is also expressed in non-sweet neurons of unknown function (Fujii et al., 2015). Finally, while Gr28bc was shown to be required for detecting and avoiding 5% saponin in flies (Sang et al., 2019), larvae showed no behavioral response to 0.1% of saponin (Choi et al., 2016)(Choi et al., 2020), a finding we independently confirmed using both concentration of saponin (data not shown).

Mapping ligand specificity for denatonium

Only Gr28ba and Gr28bc can rescue responses to denatonium when expressed in Gr28bc GRNs of ΔGr28 mutant larvae. Thus, sequence comparison between the unique N-terminal halves of the Gr28 proteins might provide insights as to possible residues important for denatonium recognition/binding, especially residues located in extracellular loops (EL) and transmembrane (TM) regions. When interrogating TM domains 1 to 4 and EL1 and EL2, only seven residues are identical between Gr28ba and Gr28bc (Table 1, Figure 7). Reducing the stringency requirement by allowing one of the remaining receptors to share the same residue, nine additional sites are identified. Interestingly, five of these sixteen amino acid residues map to four of the nine most probable binding pocket as predicted by Prankweb (https://prankweb.cz/), with three residues part of the pocket with the highest probability score (Table 2, Figure 8). Thus, targeted substitutions of single amino acids residing in predicted ligand binding pockets in other Gr28 proteins and testing such “chimera” provides a theoretical framework for practical experiments using either behavior analysis or Ca2+ imaging to precisely define the specific requirement for denatonium specificity. Additional complementary approaches can be taken to identify the critical residues in the predicted binding pockets of other members of this family once their ligands have been identified. Thus, the Gr28 proteins might serve as model to experimentally validate the predictions created by these powerful algorithms.

Amino acid alignment of the six Gr28 proteins

The sequence alignment was generated using Clustal Omega tool from ClustalW2 (https://www.ebi.ac.uk/Tools/msa/clustalo/). IL and EL indicate intracellular loop and extracellular loop, respectively. Note that the C terminal region starting at the IL3 is identical in the Gr28b proteins. Red highlighted letters indicate amino acids identical only in Gr28ba and Gr28bce. Green highlighted letters indicate amino acids conserved in Gr28ba, Gr28bc and one other Gr28 protein. An asterisk below the sequences indicates residues identical in all Gr28 proteins; a colon a strongly conserved residue (STA, NEQK, NHQK, NDEQ, QHRK, MILV, MILF, HY, FYW); and a period moderately conserved residue (CSA, ATV, SAG, STNK, STPA, SGND, SNDEQK, NDEQHK, NEQHRK, FVLIM, HFY). TM1-7 indicates helical transmembrane segment predicted by HMMTOP 2.0 software (http://www.enzim.hu/hmmtop).

Identification of candidate amino acids for ligand binding pockets of Gr28bc.

Top: Sixteen amino acids conserved in both Gr28ba and Gr28bc are mapped on 3D structure of Gr28bc using Alphafold (https://www.alphafold.ebi.ac.uk/). Red circles indicate amino acid locations of residues identical in Gr28ba and Gr28bc. Green circles indicate amino acid locations of residues identical in one additional Gr28 protein (Gr28bb or Gr28a). Bottom: Prediction of putative ligand binding pocket in of Gr28bc with highest probability using Prankweb (https://prankweb.cz/). The three resides part of this pocket and identical between Gr28ba and Gr28bc are indicated in green (V177, I218 & I222). The color displayed in the protein structure indicates a per-residue confidence score (pLDDT) between 0 and 100 as predicted by Alphafold.

Materials and methods

Key resources table

Drosophila Stocks

Flies were maintained on standard corn meal food in plastic vials under a 12 h light/dark cycle at 25°C. The w1118 strain (Bloomington Drosophila Stock Center, number 3605) was used as a wild-type control. Fly strains used: Gr28a-GAL4(SF36S for immunostaining and Figure 2B, and SF36B1 for all other experiments), Gr28ba-GAL4(NT42aC51a), Gr28bc-GAL4(NT21B1) (Thorne and Amrein, 2008); Gr28be-GAL4(Gr28a3AII) and Gr66a-GAL4 (Scott et al., 2001); ΔGr28(54B3) (Mishra et al., 2018); UAS-Gr28a, UAS-Gr28ba, UAS-Gr28bb, UAS-Gr28bc, UAS-Gr28bd and UAS-Gr28be (Ni et al., 2013); Gr43aGAL4 (Miyamoto et al., 2012); Gr66a-LexA (Thistle et al., 2012); UAS-VR1E600K (Marella et al., 2006); UAS-Kir2.1-GFP (Baines et al., 2001)(Paradis et al., 2001); lexAop-rCD2:GFP (Lai and Lee, 2006), UAS-GCaMP6m, UAS-mCD8:RFP and Gr66aex83(Bloomington Drosophila Stock Center, numbers 42748, 32220 and 35528)

Chemicals

Caffeine (Cat No. C0750), capsaicin (Cat No. M2028), denatonium benzoate (Cat No. D5765) and lobeline hydrochloride (Cat No. 141879), D-(-)-ribose (Cat No. R7500) and quinine hydrochloride dihydrate (Cat No. Q1125) were purchased from Millipore-Sigma, with a purity of > 95%. Fructose (Cat No. F1092) and agarose (Cat. No. 20-102) were purchased from Spectrum chemical and Apexbio, respectively. Sucrose (Mfr. No. 8360-06) and charcoal (Cat No. 1560-01) were purchased from Macron fine chemicals and J.T. Baker, respectively. A stock solution for capsaicin (20 mM) was prepared in 70% ethanol and stored at 4° C protected from light for up to one year. Stock solutions for bitter chemicals were prepared in Millipore Q water and stored at −20° C. Stock solutions for sugars were prepared in Millipore Q water and stored at 4° C for up to one month. A stock solution for ribose was treated with charcoal (10% of the weight of ribose used for stock solution) overnight at 4° C and sterile-filtrated (0.45 μm) to remove unrelated odor. Stock solutions were diluted to the final concentration using Millipore Q water prior to each experiment.

Immunofluorescence

Immunofluorescence of larval heads was performed based on the protocol described in Croset and colleagues (Croset et al., 2016) with minor modification. Heads of third instar larvae were dissected using microscissors in phosphate buffered saline (PBS) and immediately fixed in PBS with 4% paraformaldehyde for 1 hour at 4 °C. Heads were washed six times in washing buffer (PBS with 0.1% Triton X-100) for 20 min and blocked for 1 h in washing buffer containing 5% heat-inactivated goat serum (SouthernBiotech, Cat. No. 0060-01), followed by incubation with the primary antibodies (rabbit anti–GFP, 1:1,000 dilution; rat anti–mCD8, 1:200 dilution, ThermoFisher Scientific) at 4°C overnight. Heads were washed six times for 20 min in washing buffer and blocked in washing buffer containing 5% heat-inactivated goat serum for 1 hour. Heads were incubated with the secondary antibodies (goat anti rabbit ALEXA 488, 1:500 dilution, ThermoFisher Scientific; goat anti–rat Cy3, 1:300 dilution, Jackson Immunoresearch Laboratories Inc.) at 4°C overnight. Heads were washed six times in washing buffer for 20 min. Washes were carried out at room temperature under gentle agitation. Heads were mounted with VectaShield (Vector Lab, Cat No. H-1200) on a microscope slide and images were obtained using a Nikon A1R confocal microscope system. Adobe photoshop 2022 was used further to process images.

Larval two-choice preference assay

Two-choice preference assay of larvae was conducted as described in Mishra et al. (Mishra et al., 2013) with minor modification. Flies were placed on standard corn meal food in plastic vials and allowed to lay eggs for 24 h under a 12 h light/dark cycle at 25°C. Flies were removed from food vials and feeding-stage third-instar larvae were collected. All dishes for two-choice preference assay were prepared just prior each experiment. Petri dishes (60 x 15 mm, Falcon, Cat. No. REF353004) with two halves marked on the bottom were filled with melted plain 1% agarose or 1% agarose containing 1.75% ethanol (for capsaicin preference). After the agarose solidified, one half was removed and replaced with 1% agarose solution containing taste ligands (capsaicin or bitter compound). For each experiment, 15 larvae from food vials were briefly rinsed twice with Millipore Q water and placed along the middle separating pure and ligand containing agarose. After 16 min, photos were taken for record keeping and used to calculate larval preference indices. Larvae that crawled onto the wall of a dish or dug in the agarose were excluded. The preference index (P.I.) was calculated as follow: PI = (Ntastant - Nplain) / NTotal, whereby N is the number of larvae in the tastant sector, the plain agarose sector and the total number on the plate. Positive values indicate a preference for capsaicin or bitter compound while negative values indicate repulsion (avoidance).

Calcium imaging

Calcium imaging was performed in Gr28bc GRNs expressed in the terminal organ of feeding-stage, third instar larvae, reared as described for the larval two-choice preference assay. For each experiment, larvae from food vials were briefly rinsed twice with Millipore Q water and were mounted dorsally on a large microscope cover glass (24 x 50 mm, VWR, Cat. No. 16004-098) using double-sided scotch tape and covered with a small microscope cover glass (12CIR-1, Thermo Fisher Scientific, Cat. No. 1254580). Millipore Q water (40 µl) was applied to the tip of the larval head, and the preparation was placed on the stage of a Nikon eclipse Ti inverted microscope. Images were obtained every 500 ms, starting 15 s before application and ending 105 s after ligand application. Each recording was initiated by applying water (40μl) to set a baseline. The first ligand solution (40 μl of bitter chemical or sugar) was applied thereafter, followed by 5 washes with carrier (100 μl of water). After a 3-minute pause to allow the preparation to recalibrate, a second ligand solution (40 μl bitter chemical or sugar) was applied. To assure validity in experiments with Gr28 mutants and rescues, each recording was concluded with application of caffeine, and recordings were included only if caffeine generated a positive response. Baseline fluorescence, which was determined from the average of five frame measurements from a region next to the cell immediately before ligand application, was subtracted from the actual measurements. ΔF/F (%) = (fluorescence light intensity of the cell body – baseline/baseline) x 100. ΔF/F (max %) is the maximum value within 40 seconds after ligand application.

Generation of transgenic Gr28bc-LexA flies

To generate the Gr28bc-LexA driver, a 1.3 kb DNA fragment immediately upstream of the Gr28bc start codon was amplified from w1118flies using a forward (5′-AATCTAGGTACCCCGGCTGCTCGTCTCCCTGGATGT-3′) and a reverse (5′-CGTCAAACTAGTGACCGCTTCGTTTGAGCTTCAACC-3′) primer. Acc65I and SpeI sites included in the primer sequence (underlined) were incorporated such that the amplified fragment was amenable to directional cloning into the LexA vector CMC105 (Larsson et al., 2004). The clone chosen was confirmed by DNA sequence analysis. Transgenic flies were generated by standard P-element transformation of w1118 embryos (Rainbow Transgenic Flies Inc., Camarillo, CA).

Statistical analysis

Statistical analyses were conducted using Prism software 9.5.1 (GraphPad Software). Larval two-choice preference assay and Ca2+ imaging data were analyzed for normal distribution using D’Agostino-Pearson omnibus and Shapiro-Wilk normality tests. When groups did not meet the assumption for normal distribution, nonparametric statistics was used. For comparison between multiple groups, one-way ANOVA or Kruskal-Wallis test by ranks (nonparametric one-way ANOVA) was performed to test for difference of mean or rank distribution. As a post hoc test, Bonferroni’s or Dunn’s (nonparametric) multiple comparison tests were employed to compare two specific groups. One-way ANOVA with Bonferroni’s multiple comparison tests were used in Figure 2. Kruskal-Wallis test by ranks with Dunn’s multiple comparison tests were used in Figures 3, 4B and 5F and 5G). For comparison between two groups, Mann-Whitney U test (nonparametric t test, Figures 4A, 5D, 5E and Figure 4-figure supplement 1) with two-tailed P value was used to compare means of two groups. Sample size for larval two-choice preference assays and Ca2+ imaging experiments were based on Mishra et al (Mishra et al., 2018).

Supporting information

eLife Assessment

This valuable study focuses on the role of the Gr28 family of insect chemoreceptors. Using the Drosophila larva, the authors show that taste neurons expressing different members of this family of bitter taste receptors trigger opposite behavior - attraction and repulsion. They establish the minimal bitter taste receptor subunit composition needed in these neurons to mediate the repulsion of bitter tastants. The evidence presented is convincing, using well-validated and controlled tools and experiments.

Reviewer #1 (Public Review):

Summary:

Ahn and Amrein characterize the expression of members of the Gr28 family of gustatory receptors in taste neurons in the Drosophila melanogaster larva, define the behaviorally-relevant ligands for these receptors, and use chemogenetic experiments to show, strikingly, that different neurons have opposite behavioral responses to the chemogenetic ligand. They go on to show what neurons need to be silenced to lose responses to bitters, and very nicely show what subunits of the Gr28 bitter receptors are necessary and sufficient for responses to bitters. This is a nice piece of work, rigorously carried out, that tackles the neurons and receptors that drive innate responses to tastants in Drosophila larvae.

Strengths:

1. The chemogenetic experiments in Figure 2 are cleanly done with very clear results, and the subsetting in Figure 2B further clarifies the cellular requirements for the behavior.

2. The rescue experiments with different Gr28 subunits in the Gr28 mutant are creative and clear.

Weaknesses:

1. The authors should define early and clearly that expression of Gr28 genes studied in this paper relies not on looking at the endogenous gene but at the expression of Gal4 under the control of enhancers from these loci. The Gal4 drivers are useful and important tools, but the possibility exists that their expression is not 100% congruent with the endogenous expression of these receptors. I would not require a comprehensive validation of the lines by RNA in situ hybridization compared to the Gal4 driver lines but would recommend the disclaimer be made and that the authors are more precise in talking about the expression of the marker rather than the expression of the specific receptor gene.

2. The important chemogenetic behavioral data would benefit from a clearer presentation including a cartoon to explain what the behavior is and how it is scored. Figure 2 is the key figure in this paper and it would be helpful if the figure were re-organized to guide the non-expert reader to the key result. I recommend labeling the positive controls Gr43a as "sweet" and Gr66a as "bitter" and perhaps organize the presentation to have the negative control at the left, then Gr28ba that had no effect, then group Gr28a with Gr43a for positive valence and Gr28bc with Gr66a for negative valence. I'm not sure what the value is of showing both 0.1 mM and 0.5 mM capsaicin, the text does not explain. The experiment in Figure 2B is important but non-experts will not understand what is being done here - can the authors please provide a cartoon like those in Figure 1 showing what cells are being subjected to chemogenetics and how this differs from Figure 2A?

3. The AlphaFold ligand docking in Figure 8 is conducted with Gr28bc monomers, which are unlikely to be the in vivo relevant structure, given that the related OR/ORCO ancestor structures are tetramers. I recommend that this component of the paper either be removed entirely or that the authors redo the in silico work using the AlphaFold-Multimer package reported by Hassabis and Jumper in 2022 https://www.biorxiv.org/content/10.1101/2021.10.04.463034v2. It will be interesting to see what a tetramer structure looks like with the ligand.

Reviewer #2 (Public Review):

This study investigates how genes in the Gr28 family of gustatory receptors function in the taste system of Drosophila larvae. Gr28 genes are intriguing because they have been implicated in taste as well as other functions, such as sensing temperature and ultraviolet light. This study makes several new findings. First, the authors show that four Gr28 genes are expressed in putative taste neurons, and these neurons can be largely divided into subsets that express Gr28a versus Gr28bc. The authors then demonstrate that these two neuronal subsets drive opposing behaviors (attraction versus avoidance) when activated. The avoidance-promoting neurons respond to bitter compounds and are required for bitter avoidance, and Gr28bc and Gr28ba were specifically implicated in bitter detection in these cells. Together, these findings provide insight into the complexity of taste receptor expression and function in Drosophila, even within a single receptor subfamily.

The conclusions are well-supported by the experimental data. Strengths of the paper include the use of precise genetic tools, thorough analyses of expression patterns, carefully validated behavioral assays, and well-controlled functional imaging experiments. The role of Gr28bc neurons is more thoroughly explored than that of Gr28a neurons. However, a previous study from the same lab (Mishra et al., 2018) showed that Gr28a neurons detect RNA and ribose, which are attractive to larvae. Presumably, this is the attractive response that is being recapitulated upon artificial activation of Gr28a neurons.

I only have one technical concern: In Figure 2B, the authors do not show confirmation that using Gr66a-lexA driving lexAop-Gal80 eliminates Gal4-driven gene expression in the desired cells (cells co-expressing Gr66a and Gr28a). This is important for interpreting the behavioral experiment in order to demonstrate that the Gr28a cells mediating attraction are distinct from Gr66a/Gr28bc cells.

Reviewer #3 (Public Review):

a) Important findings

- This study confirms that Gr28 subfamily members are expressed in distinct sets of taste neurons in Drosophila larvae, supporting previous findings (e.g., Kwon et al., 2011).

- Neurons expressing different members of the Gr28 family exhibit distinct behavioral responses when chemically activated with capsaicin.

- Silencing experiments reveal that neurons expressing Gr28bc are necessary for larval avoidance of four bitter compounds, whereas neurons expressing Gr28be are necessary for avoiding lobeline and caffeine.

- Inserting either Gr28ba or Gr28bc into the GR28 mutant line restored larval avoidance of denatonium.

- Calcium imaging experiments show that Gr28ba and Gr28bc are involved in sensing denatonium, while none of the GR28 family members are involved in detecting quinine.

b) Caveats

- The authors did not acknowledge that neurons expressing members of the GR28 family also express other Gr family members, which could potentially contribute to the detection and behavioral responses to the tested bitter compounds.

- Gal4 lines from various studies exhibit varying expression patterns, highlighting the necessity for improved reagents. These findings also suggest the importance of employing different Gal4 lines for each receptor to validate the results of the current study.

- Activating or silencing neurons pertains to the function of the neurons rather than the receptors.

- Inconsistency is observed in the use of different reagents across the experiments. Specifically, all six Gal4 lines were utilized in the Chemical Activation experiments, while only two lines were employed in the silencing experiments.

- The Alphafold structure prediction is exciting but lacks conclusive evidence.

Author Response

The following is the authors’ response to the original reviews.

General comments:

To reviewer 1 and 3: The following sentences below were added at the beginning of the result section to clarify that the Gr gene expression analysis was performed using bimodal expression systems and to provide a reference that these expression profiles can generally be expected to represent endogenous Gr expression.

"Note that this and all previous Gr expression studies were performed using bimodal expression systems, mostly GAL4/UAS, whereby Gr promotors driving GAL4 are assumed to faithfully reproduce expression of the respective Gr genes. Importantly, we analyzed two or more Gr28-GAL4 insertion lines for each transgene, and at least two generated the same expression profiles (Mishra et al., 2018; Thorne and Amrein, 2008) providing evidence that the drivers reflect a fairly accurate expression profile of respective endogenous genes."

Specific comments:

Reviewer #1 (Recommendations For The Authors):

The important chemogenetic behavioral data would benefit from a clearer presentation including a cartoon to explain what the behavior is and how it is scored. Figure 2 is the key figure in this paper and it would be helpful if the figure were reorganized to guide the non-expert reader to the key result. I recommend labeling the positive controls Gr43a as "sweet" and Gr66a as "bitter" and perhaps organize the presentation to have the negative control at the left, then Gr28ba that had no effect, then group Gr28a with Gr43a for positive valence and Gr28bc with Gr66a for negative valence. I'm not sure what the value is of showing both 0.1 mM and 0.5 mM capsaicin, the text does not explain. The experiment in Figure 2B is important but non-experts will not understand what is being done here - can the authors please provide a cartoon like those in Figure 1 showing what cells are being subjected to chemogenetics and how this differs from Figure 2A?

The reviewer is correct that much can be improved, which we hope to have accomplished with the modifications in Figure 2. We re-organized it to deliver the key result to non-expert readers in an easy way. We added cartoons both explaining how the two-choice preference assays were conducted and indicating which cells express UAS-VR1. The cartoon in Figure 1E and Figure 2A are now directly relatable and should clarify what cells express VR1 (in Figure 2). Positive and negative control experiments using Gr43aGAL4 (a GAL4 knock-in; Miyamoto et al., 2013) and Gr66a-GAL4 are highlighted in the Figure and mentioned upfront in the text to make clear to what the experimental larvae can be compared. We also excluded larvae responses to 0.5 mM capsaicin.

1. The AlphaFold ligand docking in Figure 8 is conducted with Gr28bc monomers, which are unlikely to be the in vivo relevant structure, given that the related OR/ORCO ancestor structures are tetramers. I recommend that this component of the paper either be removed entirely or that the authors redo the in silico work using the AlphaFold-Multimer package reported by Hassabis and Jumper in 2022 https://www.biorxiv.org/content/10.1101/2021.10.04.463034v2. It will be interesting to see what a tetramer structure looks like with the ligand.

We tried but were able to use the recommended package. Even if it were, the problem is that we do not know the partner of Gr28b.c. And while it is not clear whether and how extensive changes in the ligand binding pockets occur when using the monomer prediciton vs a multimer package, we followed the reviewer’s suggestion and removed the modeling from the manuscript.

Minor points:

1. Line 80: I do not think it is biophysically or biochemically plausible that GRs and IRs would assemble into functional heteromeric channels and suggest that the authors either explain how that would work or remove this speculative comment.

We have removed this sentence.

1. Line 246-248: I would tone down the speculation about GR subunit composition - it's still too early days to understand the stoichiometry or the extent that any of the broadly expressed GRs is a co-receptor.

We did not indulge in the possible stoichiometry of Gr complexes, but merely mention that they are composed in general of two or more Gr subunits, for which clear genetic evidence exists: Up to three different putative bitter Gr genes are necessary to elicit responses to bitter compounds, and at least two putative sugar Gr genes are necessary to restore behavioral responses to any sweet tasting chemicals (sugars). Regardless, we have toned down the language, stating now:

“Given the multimeric nature of bitter taste receptors (Sung et al., 2017), one possibility is that the absence of a Gr subunit not required for the detection of denatonium (Gr66a) could favor formation of multimeric complexes containing Gr subunits that recognize this compound (Gr28b.a and/or Gr28b.c).”

1. Line 284: I don't think that co-expression necessarily means that GRs form heteromultimeric channels. It's equally possible that the cell controls subunit assembly to avoid mixing and matching ligand-selective subunits at will. I would tone this down - it's still speculative at this stage. We don't even know yet how this works for OR-Orco, where we do have structures. There is not yet an OR-Orco Cryo-EM structure, so we do not know what the subunit stoichiometry is.

We are not sure what the reviewer’s concern is. While direct biochemical or biophysical evidence is currently lacking, there is strong genetic evidence for heteromeric composition of Gr complexes, both from studies of bitter and sweet receptors/neurons (see response above). It is likely that intrinsic properties facilitate assembly of certain Grs within a taste receptor complex. We have refrained from any speculation about stoichiometry, though given the relatedness of Grs and Ors, it would not be far-fetched to propose that taste receptor complexes are also tetrameric in nature, which was recently proposed for a homomeric channel of the bombyx mori homolog of Gr43a, BmGr9 (Morinaga et al., 2022).

1. Line 305: the work of Emily Troemel and Cori Bargmann PMID: 9346234 should be cited in the Discussion. Theirs was the first experiment to show that valence was a feature of the neuron and not the receptor(s) it expresses.

We have now cited this work in the discussion to acknowledge this important discovery.

1. Figure 1 - the clarity of the organization of the figure could be improved for non-experts. For instance, can the key for the abbreviations be written out at the right of Figure 1A? Second, it is confusing to talk about DOG/TOG neurons "projecting" to the DO/TO - I think the authors mean dendritic innervation, not axons projecting. Maybe having a diagram that cartoons a closeup of the DOG/TOG neurons and how they innervate the cuticular structures would make this clearer. I struggled to go from the pretty staining at the left of B and C to the schematics at the right that colored in which neurons express which receptors.

We appreciate these comments regarding clarity and have amended Figure 1 and made necessary changes in the text and the Figure legend.

1. Figure 3 would benefit from a summary cartoon relating back to the cartoons in Figure 1 to summarize what neurons the authors think are necessary for bitter avoidance.

We very much appreciate this suggestion and have increased clarity by referring to the carton in Figures 1 and 2.

1. Figure 4B - the lowercase letters indicating Gr28 subunits that are being expressed under UAS control (bottom row of table "UAS-Gr28") are easily confused for the lowercase letters a, b used throughout to signify significant differences. I recommend that the authors write out the gene names in this figure to clarify the genes in the rescue experiment.

We changed the text in the Figure accordingly.

1. For non-experts it would be helpful to have a map of the Gr28 gene locus so that people understand the arrangement of the genes and how the Gal4 driver lines map onto the locus.

We have now included such a map in Figure 1B.

Reviewer #2 (Recommendations For The Authors):

1. In the title and multiple times in the text (e.g. lines 121-122), the authors make the claim that different Gr28 genes mediate opposing behaviors. At first, I was not convinced of this claim, but I now believe it may be warranted if integrating the present results with results from Mishra et al., 2018. In the present study, the authors show that different neurons drive opposing behaviors, but they did not show that the genes themselves mediate opposing behaviors. They show evidence for the role of Gr28bc and Gr28ba in aversion, but not the role of Gr28a in attraction. I was thinking that there could be other receptors in Gr28a-expressing neurons that mediate attraction. However, Mishra et al. showed that mutation of all Gr28 genes abolishes preference for RNA/ribose as well as detection of these compounds by Gr28a+ neurons of the terminal organ, an impairment that could be rescued by expressing Gr28a (although Gr28b genes seem to have similar functions), and the present study shows that the other Gr28 genes are not co-expressed with Gr28a in the terminal organ. Is this the line of reasoning that we must take to come to the conclusion in the title? If so, I don't believe it comes through clearly in the paper.

We appreciate this observation. We have modified language in the abstract and the introduction to reflect previous reports of Gr28a as an RNA/ribose receptor (Mishra et al., 2018) and its conversation across dipteran insects (Fujii et al., 2023) where we showed that appetitive behavior for RNA can be mediated via the mosquito homologs in transgenic Drosophila larvae. The reviewer is correct in that there are other appetitive neurons, namely those expressing Gr43a, which defines a set distinct from and non-overlapping with Gr28a neurons (Mishra 2018). This additional information is included in the Figure 1, summarizing expression of the Gr28 genes, Gr66a and Gr43a.

1. The Figure 6 schematic does not show Gr66a+ Gr28- cells as being connected to avoidance behavior. This seems misleading because it seems likely that these cells do promote avoidance (based on known functions of other Gr66a cells). Also, it is not clear what the red dashed line represents.

The Gr66a neurons are indeed also avoidance mediating, but it is not clear which subgroup of these neurons is necessary. Our analysis in Figure 2 using Gr28b.c driving Kir2.1 suggests that a small subset of Gr66a neurons is sufficient to mediate avoidance. It is, however, possible that other subsets not including Gr28b.c can also mediate avoidance. The figure has been modified accordingly, as has the model in Figure 7.

1. I would suggest including the description of Figures 7-8 in the Results instead of the Discussion. In Figure 8, it would be helpful to superimpose labels for the transmembrane domains and extracellular/intracellular sides to better interpret the models.

The modeling was removed from the manuscript (see response above to reviewer 1).

1. The finding that Gr66a mutants show increased denatonium and quinine avoidance (Figure 4 - figure supplement 1) seems like a non sequitur, as it does not relate to the analysis of Gr28 genes. I support the inclusion of these interesting results, but perhaps it could be stated why this experiment was conducted (e.g. as a positive control).

We have reworded this section to make clear why Gr66a mutants were tested (possibly being part of a denatonium receptor complex).

1. An introduction to the nomenclature and gene structure for the Gr28 genes would be helpful. It's not clear how they're all related, e.g. that the Gr28b genes share some exons whereas Gr28a is separate. The Results section alludes to "the high level of similarity between these receptors", and some sort of reference or quantification for this statement would be useful. I also think naming the Gr28b genes with a period (e.g. "Gr28b.c") may be more consistent with the literature.

We have added the structure of the Gr28 genes in the Figure 1B, which was also a suggestion by reviewer 1, and we have amended the naming of the genes.

1. Lines 79-80 state "some GRNs express members of both families", but no citation is provided.

As this sentence was deleted, based on a comment by reviewer 1, this point becomes mute.

1. There are several typos or grammatical mistakes that the authors may wish to correct (e.g. lines 73, 75, 91, 232, 334, 780, 788).

We appreciate the reviewer pointing these errors out to us. The mistakes were corrected.

Reviewer #3 (Recommendations For The Authors):

Silencing experiments suggest a role for Gr28bc in the avoidance of quinine (Figure 3), while imaging experiments do not support this role (Figure 5G). An explanation is needed to reconcile these findings.

The imaging experiments do support a role for Gr28b proteins in quinine detection in the specific TOG GRN used for all live imaging (Figure 5). This GRN in DGr28 larvae has a significantly lower Ca2+ responses to quinine compared to controls. However, the Ca2+ response could not be rescued to wild type levels by supplementing single Gr28b subunits, suggesting multiple Gr28b proteins are present in a quinine specific receptor complex in this GRN. Also note that Ca2+ responses of DGr28 larvae to quinine is not completely abolished, suggesting some redundancy, possible via Gr33a (Apostolopoulou et al., 2014), also supported by DGr28 larvae, which have still a robust avoidance to quinine. We are confident we have been clearer in arguing this point, both the result and especially the discussion section.

Silencing experiments specifically targeted neurons expressing Gr28bc and Gr28be (Figure 3). It is important to note why other neurons expressing different members of the Gr28 family were not included in this analysis.

Inconsistency is observed in the use of different reagents across the experiments. Specifically, all six Gal4 lines were utilized in the Chemical Activation experiments, while only two lines were employed in the silencing experiments.

The silencing experiments asked the specific questions as to what neurons are necessary for avoidance of bitter chemicals. Gr28a-GAL4 and Gr28b.a-GAL4 neurons were omitted because the former mediate feeding preference and not avoidance, and the latter is expressed in the same neurons as Gr28b.e (Figure 1). The remaining two Gr28b genes, Gr28b.b-GAL4 and Gr28b.d-GAL4 are not expressed in the larval taste system (Mishra et al., 2018) as we stated in the introduction/result section, and they were therefore not included in the chemogenetic or Kir2.1 inactivation experiments. We included these genes in rescue experiments, simply to test whether or not they can restore function for sensing denatonium.

As for the chemogenetic activation experiments: two of the GAL4 lines are controls (Gr66a-GAL4 and Gr43GAL4), that were needed to show what can be expected from these experiments.

The authors did not acknowledge that neurons expressing members of the GR28 family also express other Gr family members, which could potentially contribute to the detection and behavioral responses to the tested bitter compounds.

We believe we did, but we have made that much more explicit in the revised manuscript.

Gal4 lines from various studies exhibit varying expression patterns, highlighting the necessity for improved reagents. These findings also suggest the importance of employing different Gal4 lines for each receptor to validate the results of the current study.

See response at the beginning of our rebuttal.

Activating or silencing neurons pertains to the function of the neurons rather than the receptors.

We agree and nothing in the manuscript states otherwise.

Version 2

Abstract

Abstract

Most animals have functionally distinct populations of taste cells, expressing receptors that are tuned to compounds of different valence. This organizational feature allows for discrimination between chemicals associated with specific taste modalities and facilitates differentiating between unadulterated foods and foods contaminated with toxic substances. In the fruit fly D. melanogaster, primary sensory neurons express taste receptors that are tuned to distinct groups of chemicals, thereby activating neural ensembles that elicit either feeding or avoidance behavior. Members of a family of ligand gated receptor channels, the Gustatory receptors (Grs), play a central role in these behaviors. In general, closely related, evolutionarily conserved Gr proteins are co-expressed in the same type of taste neurons, tuned to chemically related compounds, and therefore triggering the same behavioral response. Here, we report that members of the Gr28 subfamily are expressed in largely non-overlapping sets of taste neurons in Drosophila larvae, detect chemicals of different valence and trigger opposing feeding behaviors. We determined the intrinsic properties of Gr28 neurons by expressing the mammalian Vanilloid Receptor (VR1), which is activated by capsaicin, a chemical to which wildtype Drosophila larvae do not respond. When VR1 is expressed in Gr28a neurons, larvae become attracted to capsaicin, consistent with reports showing that Gr28a itself encodes a receptor for nutritious RNA. In contrast, expression of VR1 in two pairs of Gr28b.c neurons triggers avoidance to capsaicin. Moreover, neuronal inactivation experiments show that the Gr28b.c neurons are necessary for avoidance of several bitter compounds. Lastly, behavioral experiments of Gr28 deficient larvae and live Ca2+ imaging studies of Gr28b.c neurons revealed that denatonium benzoate, a synthetic bitter compound that shares structural similarities with natural bitter chemicals, is a ligand for a receptor complex containing a Gr28b.c or Gr28b.a subunit. Thus, the Gr28 proteins, which have been evolutionarily conserved over 260 million years in insects, represent the first taste receptor subfamily in which specific members mediate behavior with opposite valence.

Introduction

Meaningful animal behavior is established through cooperative engagement of multiple sensory systems. In many insects, the chemosensory system plays a central role in such integration processes. The fruit fly Drosophila melanogaster has served as the primary insect model system for elucidating the molecular basis and neural circuitry of both olfaction and taste, by virtue of the vast genetic resources, amenability to both neurophysiological recording and live imaging, and simple yet powerful behavioral assays, allowing investigators to link genes to chemosensory behavior and neural activity (Montell, 2021).

The Drosophila gustatory system is characterized by several insect specific anatomical and molecular features. For example, taste cells are primary sensory neurons, with dendritic processes that express taste receptors, while long axons project and convey taste information directly to the brain. In adult flies, these neurons, referred to as Gustatory Receptor Neurons (GRNs) are distributed across several appendages, such as labial palps, legs and presumably the antennae. Additionally, some GRNs reside internally, arranged in cell clusters along the pharynx (Amrein, 2016; Joseph and Carlson, 2015; Scott, 2018). Likewise, Drosophila larvae have numerous structures located both on the head surface as well as internally along the larval pharynx, where chemical compounds are assessed during their passage towards the digestive system (Figure 1A)(Apostolopoulou et al., 2015; Kwon et al., 2011; Rist and Thum, 2017).

Expression of the Gr28 genes in the larval sensory organs.

A) Schematic representation of the larval chemosensory system. Three external sensory organs (DO, TO and VO) hold collectively the dendritic extensions of neuronal cell bodies in the respective ganglia (DOG, TOG and VOG). Three clusters of sensory neurons (DPS, VPS and PPS) reside along the pharynx. The antennal nerve (AN) connects the DOG neurons to the subesophageal zone (SEZ). The TOG and VOG neurons project along the maxillary nerve (MN) to the SEZ. The DPS and PPS neurons project axons to the SEZ via the labral nerve (LRN). The VPS neurons project to the SEZ through the labial nerve (LBN). Note that olfactory neurons are omitted in the schematic here and in E (below).

B) Structure of the Gr28 locus. The six Gr28 genes are clustered within 10 kilobases. The five Gr28b genes are transcribed from distinct promotors, with first unique exons that are spliced to common second and third exons. Gr28a is downstream of the Gr28b genes. The Gr28 mutation (ΔGr28) used in this study lacks the shared common exons of the Gr28b genes and the entire Gr28a gene.

C) Expression of the Gr28 genes in the external sensory organs. Note that images show only one of the bilaterally symmetrical organs. Co-expression between Gr66a (a marker for bitter taste GRNs) and different Gr28 genes (in “Merge” panel) was assessed using GAL4 and LexA drivers for Gr28 genes and Gr66a, respectively. For Gr28b.a, Gr28b.c and Gr28b.e co-expression was observed in each case. However, Gr66a and Gr28a are expressed in different GRNs. Asterisks refer to a GRN expressing both Gr66a and the indicated Gr28 gene. Scale bars are 5 μm.

D) Expression of the Gr28 genes in internal sensory organs. Note that the bilaterally symmetrical halves of the DPS/VPS are physically close to each other, and the images includes both halves, while the image of the PPS shows only one side of the bilaterally symmetrical organ. Gr28b.c, but none of the other Gr28b genes, is co-expressed with Gr66a in the DPS/VPS. In the PPS, none of the Gr28b genes is found, while Gr28a and Gr66a are partially co-expressed, but each gene is also expressed exclusively in a subset of GRNs. Asterisks refer to a GRN expressing both Gr66a and the indicated Gr28 gene. Scale bars are 5 μm.

E) Expression summary: Only relevant neurons in one of the paired taste organs, are shown with total number of neurons indicated in parenthesis. The cartoon summarizes the immunostainings shown in C) and D). The VOG is not shown as none of the Gr28-GAL4 drivers is expressed there. Immunostaining was performed on whole-mount preparations from larvae heads of the following genotypes: UAS-mCD8:RFP lexAop-rCD2:GFP;Gr66a-LexA/Gr28a-GAL4, UAS-mCD8:RFP lexAop-rCD2:GFP;Gr66a-LexA/Gr28b.c-GAL4, UAS-mCD8:RFP lexAop-rCD2:GFP;Gr66a-LexA/Gr28b.e-GAL4 and UAS-mCD8:RFP lexAop-rCD2:GFP;Gr66a-LexA/+; Gr28b.a-GAL4/+. The Gr43aGAL4 GRNs, which do not overlap with Gr28a-GAL4 neurons (Mishra et al., 2018) are shown for reference to experiments described in Figures 2 and 7.

Fruit flies, like most insects, employ taste receptors encoded by two major gene families, the Gustatory receptor (Gr) and the Ionotropic Receptor (IR) genes, to sense soluble chemicals, such as appetitive food compounds, noxious and toxic chemicals as well as pheromones. Both Gr- and IR- based receptors are thought to form complexes composed of several, and generally different, subunits. Composition of only a few Gr-based taste receptors for sugars and bitter compounds is known (Jiao et al., 2008; Yavuz et al., 2014; Shim et al., 2015; Sung et al., 2017) (Fujii et al., unpublished), and two Gr proteins, Gr21a and Gr63a, are co-expressed in a small subset of olfactory neurons where they form a receptor complex for gaseous carbon dioxide (Jones et al., 2007; Kwon et al., 2007).

The Gr genes represent the largest taste receptor gene family in insects. In D. melanogaster, it is comprised of 60 genes predicted to encode 68 proteins, expression of which has been extensively described (Dahanukar et al., 2007; Dunipace et al., 2001; Fujii et al., 2015; Scott et al., 2001; Weiss et al., 2011). Several Gr genes have been functionally characterized in adult flies using genetic mutations combined with either electrophysiological recordings, Ca2+ imaging studies and behavioral analyses (Amrein, 2016; Montell, 2021), but only a few have been studied in larvae (Apostolopoulou et al., 2016a; Choi et al., 2016, 2020; Mishra et al., 2013, 2018).

The Gr28 gene subfamily is of particular interest for a number of reasons: First, it is one of the most conserved Gr subfamilies, homologs of which can be found across all insect families and even more distant arthropods (Eyun et al., 2017; Fujii et al., 2023; Suzuki et al., 2018). The six Gr28 genes are tightly clustered, with the five Gr28b genes transcribed from distinct promotors and unique exons spliced to shared 2nd and 3rd exons, while Gr28a is a separate transcription unit (Figure 1B). Overall conservation between the Gr28 proteins is high (equal or more than 50% similarity), characteristic for genes generated through recent gene duplication events. Second, the Gr28 genes are expressed not only in the gustatory system of larvae and adult flies, but also in many other organs, especially the central nervous system (CNS) and non-chemosensory neurons of the peripheral nervous system (PNS), suggesting that they have functions beyond gustation and are important to sense chemical signals unrelated to food (Mishra et al., 2018; Thorne and Amrein, 2008). Notably, evidence for such roles has been reported before any direct link to gustatory perception was discovered. Ni and colleagues showed that Gr28b.d is essential for high temperature avoidance in flies (Ni et al., 2013), while Xiang and collaborators found that Gr28 mutant larvae were deficient in UV light avoidance (Xiang et al., 2010). Third, the only known gustatory function for any Gr28 protein thus far is sensing of RNA and ribose by Gr28a, mediated by Gr28a-GAL4 GRNs (Mishra et al., 2018). Remarkably, sensing RNA and ribose is an appetitive taste quality found in other dipteran insects that diverged from Drosophila about 260 million years ago, including flesh flies and mosquitoes, and we showed that Gr28 homologs from A. aegypti and A. gambiae can restore RNA and ribose preference in Drosophila Gr28 mutant larvae when expressed in Gr28a-GAL4 GRNs (Fujii et al., 2023).

Here, we present a detailed expression analysis and functional characterization of the Gr28 genes and the respective GRNs in Drosophila larvae. In addition to Gr28a, three of the five Gr28b genes (Gr28b.a, Gr28b.c and Gr28b.e) are also expressed in the larval taste system. Interestingly, GRNs expressing Gr28a and the three Gr28b genes represent functionally distinct neuronal ensembles, with minimal expression overlap in a single pair of neurons expressing both Gr28a and Gr28b.c. When the mammalian Vanilloid Receptor 1 (VR1) is expressed under the control of specific GAL4 drivers, we found that Gr28a-GAL4 and Gr28b.c-GAL4 neurons mediate opposing taste behavior in the presence of capsaicin, the ligand for VR1. Specifically, Gr28a-GAL4/UAS-VR1 larvae show strong attraction for capsaicin, while Gr28b.c-GAL4/UAS-VR1 larvae show strong avoidance of capsaicin. Neuronal inactivation experiments reveal that the Gr28b.c GRNs are necessary to sense bitter compounds, such as denatonium, quinine, lobeline, and caffeine. Moreover, Ca2+ responses of Gr28b.c GRNs to denatonium and quinine are significantly reduced and avoidance behavior of these two chemicals is diminished in Gr28 mutant larvae, whereas Ca2+ responses and avoidance behavior were not affected when challenged with lobeline and caffeine. This implies that Gr28b proteins are subunits of receptor complexes sensing a subgroup of bitter tasting compounds. In summary, the Gr28 genes encode related Gr proteins mediating both positive and negative valence.

Results

Expression of the Gr28 genes in the larval taste organs

The peripheral chemosensory system of the larvae is subdivided into bilaterally symmetrical, ‘external’ and ‘internal’ taste organs (Stocker, 2008). The three external organs reside near the tip of the head and are organized as paired ganglia, the Dorsal, Terminal and Ventral Organ Ganglia (DOGs, TOGs and VOGs) that house the GRN cell bodies with dendritic extensions in respective sensory organs (DO, TO and VO) at the head surface, while carrying information via their axons to the subesophageal zone (SEZ) in the brain (Figure 1A). The DOG harbors 21 olfactory neurons and 18 presumptive GRNs. For clarity, GRN numbers refer to neurons in one of the two, bilaterally symmetrical taste organs. The GRNs located in the DOG fall into two distinct groups, based on their dendritic extensions: 11 presumptive GRNs, four of which were shown to express Gr genes (Apostolopoulou et al., 2015), send dendrites to the base of the DO, while 7 neurons extend dendrites to the TO (the dorsolateral group, Figure 1A)(Kwon et al., 2011). The TOG contains 30 neurons, with dendrites located in the TO (the distal group). The internal taste structures, referred to as the Dorsal/Ventral Pharyngeal Sense organ (DPS/VPS, 33 neurons) and the Posterior Pharyngeal Sense organs (PPS, 6 neurons) are located along the pharynx and sense chemicals as they are moved towards the digestive system. We note that not all these neurons are confirmed GRNs, either by function or expression of Gr or IR genes, albeit based on their location and anatomy, most are thought to be GRNs (Rist and Thum, 2017; Sánchez-Alcañiz et al., 2018; Stewart et al., 2015).

Our initial expression analysis of the Gr28 genes revealed that four of the six Gr28 genes (Gr28a, Gr28b.a, Gr28b.c and Gr28b.e) were expressed in larval taste organs, in addition to cells in the gut, the brain and non-chemosensory cells of the larvae (Mishra et al., 2018; Thorne and Amrein, 2008). This and all previous Gr expression studies were performed using bimodal expression systems, mostly GAL4/UAS, whereby Gr promotors driving GAL4 are assumed to faithfully reproduce expression of the respective Gr genes. Importantly, we analyzed two or four Gr28-GAL4 insertion lines for each transgene, and at least two generated the same expression pattern (Mishra et al., 2018; Thorne and Amrein, 2008) providing evidence that the drivers reflect a fairly accurate expression profile of respective endogenous genes. To further delineate the putative chemosensory roles of these genes, we performed a more detailed co-expression analysis between the Gr28 genes and the bitter receptor gene Gr66a by combining the GAL4/UAS (labeling Gr28 expressing neurons) with the LexA/lexAop (marking Gr66a neurons) system (Figure 1C and 1D). In the TOG, we found expression of all four Gr28 genes, along with that of Gr66a, which was expressed in 3 or 4 neurons (this number is slightly smaller than the 6 neurons previously reported by (Kwon et al., 2011). GAL4 drivers for Gr28b.a, Gr28b.c and Gr28b.e are co-expressed in a single neuron with Gr66a-LexA (Figure 1C). In contrast, Gr28a-GAL4 is expressed in a distinct TOG neuron than Gr66a-LexA (Figure 1C) and the Gr28b genes, an observation we independently confirmed using a Gr28b.c-LexA driver (Figure 1-figure supplement 1). In the DOG, we find only a single Gr28a-GAL4 neuron, while none of the Gr28b genes, or Gr66a, is expressed there (Figure 1C). None of the Gr28-GAL4 drivers was expressed in the VOG. In the internal sensory organs, Gr66a-LexA is found in one GRN in the DPS/VPS, where it is co-expressed with Gr28b.c-GAL4 as well as Gr28a-GAL4 (Figure 1D). In the PPS, none of the Gr28b genes is expressed, but Gr28a-GAL4 is found in two neurons, one of which also expresses Gr66a-LexA (Figure 1D). In summary, the internal sensory organs can be subdivided in three distinct groups (Figure 1E): Gr66a/Gr28b.c/Gr28a positive neurons, Gr66a/Gr28a positive neurons and Gr28aonly neurons, which contrasts the external sensory organs where Gr28a and the Gr28b genes are expressed in a mutually exclusive fashion.

Subsets of Gr28 neurons mediate opposing feeding behaviors

We previously showed that at least one of the six Gr28 genes is necessary for feeding attraction to RNA, ribonucleosides and ribose using a well-established two choice feeding assay (Mishra et al., 2018)(Figure 2A). Specifically, larvae homozygous mutant for Gr28 (ΔGr28, a deletion of the entire Gr28a gene and more than half of the coding region of all Gr28b genes, Figure 1B), lose their ability to sense these compounds, a phenotype that is restored when single UAS-Gr28 reporter transgenes are expressed in Gr28a-GAL4 neurons. The largely non-overlapping expression of Gr28a and the Gr28b genes suggests that respective neurons represent functionally distinct entities. To investigate this possibility, we took advantage of the mammalian Vanilloid Receptor 1 (VR1), a TRP channel that is activated by capsaicin (Caterina et al., 1997). Drosophila have no VR1 like-gene in their genome and do not respond to capsaicin behaviorally, but flies are attracted to this chemical when a modified VR1 gene (VR1E600K) is expressed in sweet taste neurons (Marella et al., 2006). Thus, we expressed VR1E600K (henceforth referred to UAS-VR1) under the control of the four Gr28-GAL4 drivers (Gr28a-GAL4, Gr28b.a-GAL4, Gr28b.c-GAL4 and Gr28b.e-GAL4) in larvae and tested their response to capsaicin using the two-choice preference assay (Figure 2A). Additionally, we expressed UAS-VR1 in bitter neurons as well as appetitive fructose sensing neurons using respective GAL4 drivers (Gr66a-GAL4 and Gr43aGAL4) (Mishra et al., 2013; Scott et al., 2001), which served as control larvae (note that fructose sensing Gr43aGAL4 neurons are distinct from Gr28a-GAL4 neurons (Mishra et al., 2018)). Just like adult flies, w1118control larvae lacking either a UAS-VR1 reporter, a Gr-GAL4 driver, or both, were unresponsive to 0.1 mM capsaicin, while the positive control larvae expressing UAS-VR1 in either fructose sensing GRNs or bitter taste GRNs showed robust attraction to or avoidance of capsaicin (Figure 2B). Consistent with previous findings, which identified Gr28a GRNs as appetitive neurons (Mishra et al., 2018), larvae expressing UAS-VR1 under the control of Gr28a-GAL4 showed strong appetitive responses to capsaicin (Figure 2B). In contrast, larvae expressing UAS-VR1 under the control of Gr28b.c-GAL4 showed robust avoidance behavior to capsaicin, while expression in the single TOG GRN using either Gr28b.a-GAL4 or Gr28b.e-GAL4 (also expressing Gr28b.c; Figure 1C) caused neither attraction to nor avoidance of capsaicin.

Intrinsic valence of different Gr28 GRNs

A) Behavioral two-choice preference assay to quantify appetitive and avoidance of larvae for chemical ligands.

15 feeding stage, 3rd instar larvae are placed along the demarcation of a 1% agarose containing dish (35 mm), with one side plain, and the other side containing ligand. The preference Index (P.I.; see Material and Methods) is calculated after counting location of larvae after 16 minutes.

B) Gr28a GRNs mediate capsaicin preference, while Gr28b.c GRNs elicit capsaicin avoidance in larvae. GRNs expressing the capsaicin receptor VR1 (brown panels) are marked with a red crown in diagrams below. w1118, reporter gene only (UAS-VR1/+) and respective GAL4 driver (Gr-GAL4/+) larvae serve as negative controls (white panels) and show neither preference for nor avoidance to 0.1 mM capsaicin. Experimental larvae expressing VR1 in Gr28-GAL4 neurons are shown in blue. Positive control larvae (purple panels) expressing VR1 in bitter taste GRNs (Gr66a-Gal4) or appetitive, sweet GRNs (Gr43aGAL4) show expected avoidance to or preference for capsaicin. Experimental larvae (blue panels) expressing VR1 in Gr28a-GAL4 GRNs display robust preference for capsaicin, which is still observed when expression is further restricted to Gr28aonly GRNs (Gr66a-LexA/UAS-VR1; Gr28a-GAL4/lexAop-GAL80; right panel). In contrast, when VR1 is expressed in Gr28b.c-GAL4 GRNs, larvae strongly avoid capsaicin. Neither avoidance nor preference was observed when VR1 is expressed in the Gr28b.a-GAL4 or Gr28b.e-GAL4 GRNs. Each bar represents the mean ± SEM of P.I. (n = 10-22 assays). The taste behavior of Gr-GAL4 > UAS-VR1 larvae is compared to 3 controls (w1118, UAS-VR1/+ and Gr-GAL4/+) using one-way ANOVA with Bonferroni’s multiple comparison tests (p < 0.05), whereby different letters indicate a statistically significant difference. Dashed lines delineate groups for ANOVA. All control and experimental larvae are in the w1118 background, carry one copy of the indicated transgene(s) and were generated from crosses of respective strains listed in the Material and Methods.

Overlap between Gr28a and Gr66a in internal GRNs raises the question about their contribution to appetitive and/or avoidance behavior. We deemed it unlikely that these neurons were critical for appetitive behavior, since the one pair located in the DPS/VPS (expressing also Gr28b.c) is necessary for capsaicin avoidance (see above). Indeed, when VR1 was suppressed in all Gr66a/Gr28b.c neurons by means of the GAL4 suppressor GAL80, these larvae, expressing VR1 in Gr28aonly GRNs remained strongly attracted to capsaicin (Figure 2B last panel; for effective suppression, see Figure 2- figure supplement 1). Taken together, this analysis indicates that Gr28a-GAL4 is expressed in appetitive inducing neurons, while Gr28b.c-GAL4 neurons mediate avoidance behavior.

The capsaicin experiments above suggest that the two larval Gr28b.c-GAL4 GRNs mediate negative valence. Previous reports have shown that larvae avoid many bitter compounds (Apostolopoulou et al., 2016b, 2015; Choi et al., 2020, 2016; van Giesen et al., 2016), and thus, we expected that eliminating activity of Gr28b.c-GAL4 GRNs would result in loss of avoidance behavior to at least some bitter chemicals. Indeed, when Gr28b.c-GAL4 GRN activity was blocked using the inward rectifying potassium channel Kir2.1 (Baines et al., 2001; Paradis et al., 2001), larvae no longer avoided the four tested bitter compounds denatonium, quinine, lobeline and caffeine (Figure 3). A compound specific avoidance phenotype to lobeline and caffeine was observed when the Gr28b.e-GAL4 GRNs in the TOG was inactivated. These data, together with the capsaicin experiments, suggest that two pairs of GRNs, one in the TOG and one in the DPS/VPS are necessary and sufficient for avoidance of these four bitter tasting chemicals.

The Gr28b.c neurons mediate avoidance behavior to bitter compounds

Inactivation of Gr28b.c GRNs (Gr28b.c-GAL4/UAS-Kir2.1) elicits significantly reduced avoidance of larvae to all four bitter compounds tested - denatonium, quinine, lobeline and caffeine - while control larvae, carrying either the driver or the reporter only, showed strong avoidance of these compounds. In contrast, larvae with inactivated Gr28b.e GRNs (Gr28b.e-GAL4/UAS-Kir2.1) still avoid denatonium and quinine (top), but no longer avoid lobeline and caffeine (bottom). Each bar represents the mean ± SEM of P.I. (n = 11- 20 assays). The taste behavior of Gr28b.c-GAL4; UAS-Kir2.1 and Gr28b.e-GAL4; UAS-Kir2.1 larvae is compared to 2 controls (UAS-Kir2.1/+ and Gr28b.c -GAL4/+ or Gr28b.e-GAL4/+) using Kruskal-Wallis test by ranks with Dunn’s multiple comparison tests (p < 0.05). Bars with different letters are significantly different. Dashed lines delineate groups for ANOVA. Fly genotypes: w1118; UAS-Kir2.1/+ (light gray), w1118; Gr28b.c-GAL4/+ (dark gray), w1118; Gr28b.c-GAL4/UAS-Kir2.1 (white), w1118; Gr28b.e-GAL4/+ (dark gray), w1118; Gr28b.e-GAL4/UAS-Kir2.1 (white).

Gr28b.c and Gr28b.a are subunits of a taste receptor complex for denatonium

We next examined whether any of the Gr28b proteins is part of a taste receptor complex detecting any of these bitter chemicals (Figure 4). Surprisingly, only avoidance of denatonium was affected in larvae lacking the Gr28 gene cluster (ΔGr28; Figure 1B), while avoidance to quinine was somewhat reduced, albeit not significantly, and avoidance to lobeline or caffeine remained robust. In fact, avoidance to caffeine increased modestly, but significantly (Figure 4A). We then examined whether any of the Gr28 proteins was sufficient to restore denatonium avoidance by expressing individual Gr28 genes under the control of the Gr28b.c-GAL4 driver. Indeed, either Gr28b.a or Gr28b.c expression led to a full recovery of denatonium avoidance, while expression of any other Gr28b gene, or Gr28a, failed to do so (Figure 4B). This observation suggests that despite the high level of similarity between these receptors, recognition of denatonium is dependent on specific structural features present in Gr28b.a and Gr28b.c, but not in any of the other Gr28 proteins.

Role of individual Gr28 genes in bitter taste avoidance

A) Gr28 genes are required for sensing denatonium. Wild-type (w1118) larvae, but not Gr28 mutant larvae (w1118;ΔGr28/ΔGr28), strongly avoid denatonium. In contrast, w1118;ΔGr28/ΔGr28 larvae do not show significantly reduced avoidance to quinine, lobeline and caffeine. Each bar represents the mean ± SEM of P.I. (n = 12- 22 assays). Asterisks indicate a significant difference between w1118; ΔGr28/ΔGr28 and w1118 larvae (Two-tailed, Mann-Whitney U test, **** p < 0.0001, ns: not significant).

B) Single Gr28b genes can rescue avoidance response to denatonium when expressed in Gr28b.c neurons of Gr28 mutant larvae. The behavior of w1118;ΔGr28/ΔGr28 larvae expressing UAS-Gr28 transgenes under control of the Gr28b.c-GAL4 driver was compared to Gr28+/+ controls (w1118, black bar), and three Gr28 mutant controls (ΔGr28, ΔGr28 plus driver, and ΔGr28 plus respective UAS-Gr28 reporter, white bar) using Kruskal-Wallis test by ranks with Dunn’s multiple comparison tests (p < 0.05). Each bar represents the mean ± SEM of P.I. (n = 11- 22 assay). Bars with different letters are significantly different. Dashed lines delineate groups for ANOVA. Fly genotypes: wild-type: w1118 (black), mutants: w1118;ΔGr28/ΔGr28, w1118; ΔGr28/ΔGr28 Gr28b.c-GAL4, w1118;ΔGr28/ΔGr28; UAS-Gr28 (indicated Gr28 genes)/+ and w1118; ΔGr28/ΔGr28 UAS-GCaMP6m; UAS-Gr28 (for Gr28b.b or Gr28b.c genes)/+ (white), rescues: w1118; ΔGr28/ΔGr28 Gr28b.c-GAL4; UAS-Gr28 (indicated Gr28 genes)/+ and w1118; ΔGr28 UAS-GCaMP6m/ΔGr28 Gr28b.c-GAL4; UAS-Gr28 (for Gr28b.b or Gr28b.c genes)/+ (gray).

Since Gr66a-LexA is co-expressed in all Gr28b-GAL4 expressing GRNs, we wondered whether Gr66a is a component of the denatonium receptor. Previous work had established that Gr66a is required for caffeine avoidance, in both larvae and adult flies (Apostolopoulou et al., 2016b; Lee et al., 2009; Moon et al., 2006), which we confirmed (Figure 4-figure supplement 1). Surprisingly, avoidance of denatonium and quinine was not diminished, but increased significantly (Figure 4-figure supplement 1). Given the multimeric nature of bitter taste receptors (Sung et al., 2017), one possibility is that the absence of a Gr subunit not required for the detection of denatonium (Gr66a) could favor formation of multimeric complexes containing Gr subunits that recognize this compound (such as Gr28b.a and/or Gr28b.c).

Finally, we investigated neuronal responses in larvae expressing the Ca2+ indicator GCaMP6m in Gr28b.c-GAL4 GRNs (Figure 5). We developed a whole animal imaging preparation, whereby larvae were placed in an “imaging chamber” to minimize head movements (Figure 5A), and visualized neural activity of the Gr28b.c-GAL4 GRN in the TOG in real time (Chen et al., 2013) upon exposure to the four bitter compounds, as well as sucrose, ribose and fructose (Figure 5B to 5D). All bitter compounds elicited rapid Ca2+ increases in Gr28b.c-GAL4 GRNs, while none of the sugars did (Figure 5C and 5D). When neural activity was recorded in Gr28b.c-GAL4 GRNs of ΔGr28 homozygous mutant larvae (Figure 5E), Ca2+ responses to denatonium and quinine were severely reduced, while responses to both caffeine and lobeline were not affected. Re-expression of either Gr28b.a or Gr28b.c, but not Gr28b.b, Gr28b.d, Gr28b.e or Gr28a rescued Ca2+ response to denatonium, but not to quinine (Figure 5F and 5G). Together, these experiments identified Gr28b.c and Gr28b.a as redundant subunits of a denatonium receptor complex, a complex that does not require Gr66a or any of the other Gr28b subunits.

Cellular Ca2+ responses of Gr28b.c GRNs to select bitter compounds requires Gr28b.a or Gr28b.c.

A) Diagram of Ca2+ imaging experimental set up.

B) Representative still images of Ca2+ response in the Gr28b.c expressing GRN of the TOG. Ca2+ responses of the Gr28b.c GRNs upon stimulation with indicated ligands. ΔF indicates the changes in fluorescence light intensity of the cell body after ligand application.

C and D) Representative traces (C) and quantified Ca2+ responses (D) of the Gr28b.c GRNs after stimulation with indicated ligands. Fly genotype: w1118; Gr28b.c-GAL4/UAS-GCaMP6m. Each bar represents the mean ± SEM of Ca2+ imaging with 12 – 16 larvae. Asterisks indicate a significant difference between carrier (water) and indicated ligands (Two-tailed, Mann-Whitney U test, *** p < 0.001, ns: not significant).

E) Neurons of larvae lacking the Gr28 genes exhibit significantly reduced responses to denatonium and quinine. Gr28b.c expressing GRNs in the TOG of Gr28 mutant larvae (ΔGr28) have significantly reduced Ca2+ responses to denatonium and quinine but not to lobeline or caffeine when compared to Gr28b.c expressing GRNs of wild type controls. Larvae genotypes: Gr28+ control (black bar): w1118; Gr28b.c-GAL4/UAS-GCaMP6m. ΔGr28 control (white bar): w1118; ΔGr28 Gr28b.c-GAL4/ΔGr28 UAS-GCaMP6m. Each bar represents the mean ± SEM with 13 – 16 larvae. Asterisks indicate a significant difference between Gr28+ control and ΔGr28 larvae (Two-tailed, Mann-Whitney U test, *** p < 0.001, ** p < 0.01 ns: not significant).

F and G) Gr28b.c or Gr28b.a transgenes rescue denatonium responses in GRNs of ΔGr28 larvae. Expression of Gr28b.c or Gr28b.a is under control of Gr28b.c-GAL4 restores responses to denatonium, but not quinine in TOG GRNs of ΔGr28 larvae. Each bar represents the mean ± SEM of Ca2+ imaging with 12 – 17 larvae. The Ca2+ responses of Gr28 mutant larvae expressing UAS-Gr28 transgenes under Gr28b.c-GAL4 driver is compared to Gr28+ (black) and ΔGr28 (white) controls using Kruskal-Wallis test by ranks with Dunn’s multiple comparison tests (p < 0.05). Bars with different letters are significantly different. Dashed lines delineate groups for ANOVA. Fly genotypes: Gr28+ control (black bar): w1118; Gr28b.c-GAL4/UAS-GCaMP6m. ΔGr28 control (white bar): w1118; ΔGr28 Gr28b.c-GAL4/ΔGr28 UAS-GCaMP6m. Gr28 rescues (gray bar): w1118; ΔGr28 Gr28b.c-GAL4/ΔGr28 UAS-GCaMP6m; UAS-Gr28 (indicated Gr28 genes)/+. Concentration of ligands was 100 mM for sugars, 50 mM for caffeine and 5 mM for denatonium, quinine and lobeline.

Discussion

The Gr28 receptors comprise one of the few Gr subfamilies conserved across diverse insect species (Agnihotri et al., 2016; Engsontia and Satasook, 2021; Yu et al., 2023). Yet, they were the least characterized when compared to other conserved subfamilies, such as the sugar receptors (Gr5a, Gr61a and Gr64a-f), the carbon dioxide receptors Gr21a and Gr63a or the bitter taste receptors. The only ligands associated with the Gr28 proteins were ribonucleosides and RNA, which are appetitive nutrients essential for larvae and detected by Gr28a neurons (Mishra et al., 2018). Indeed, RNA has been found to be an appetitive taste ligand across many dipteran insects, including mosquitoes, and we showed that Gr28 homologs of both A. aegypti and A. gambiae can rescue the preference for RNA and ribose when expressed in Gr28a neurons of ΔGr28 mutant larvae (Fujii et al., 2023).

Previous studies in adult Drosophila have shown that members of conserved Gr protein families such as the carbon dioxide receptors (Gr21a and Gr63a) (Jones et al., 2007; Kwon et al., 2007) or the receptors for sweet taste encoded by the eight sugar Gr genes (Fujii et al., 2015) are largely co-expressed in one type of neuron in the fly’s taste organs. For example, with the exception of Gr5a (see below), sugar Gr genes are only expressed in a single GRN (the “sweet’ neuron) of each taste sensilla, and activation of these “sweet” neurons by sugars requires the function of at least two of the 8 sugar Gr genes (Dahanukar et al., 2007; Yavuz et al., 2014)(Fujii et al., unpublished). Similarly, the approximately 33 putative bitter taste receptors, which comprise several small conserved subfamilies (Robertson et al., 2003), as well as individual Gr genes with little overall similarity to one another, are partially co-expressed in a the bitter GRN of each taste sensilla (Weiss et al., 2011). Molecular genetic studies combined with electrophysiological recordings have shown that at least three different Gr subunits are required to constitute functional receptor complexes that can sense a bitter compound (Shim et al., 2015). We note that two rare exceptions to the heteromeric nature of taste receptor complexes exist, namely the RNA receptor Gr28a and the fructose receptor Gr43, which have been proposed to function as homomultimeric complexes (Mishra et al., 2018, 2013). Cryo-EM structural analysis of the conserved insect Olfactory receptor co-receptor (ORCO) suggests that insect odorant receptors form tetramers (Butterwick et al., 2018), and biochemical characterization and comparative modeling of BmGr9, the Bombyx mori homolog of the Drosophila Gr43a fructose receptor, supports such structures for Gr proteins as well (Morinaga et al., 2022).

Distinct functions are mediated by small set of GRNs expressing specific Gr28 subunits

Our expression analysis of the bitter taste receptor gene Gr66a and the Gr28 genes in larvae is consistent with earlier studies, despite some small variation in neuron number (Choi et al., 2016; Kwon et al., 2011), which is likely due to the use of different GAL4 driver lines and/or variability in expression levels. Importantly, all Gr28b genes are co-expressed with the bitter taste receptor Gr66a (Figure1) and probably several other putative bitter Gr genes (Kwon et al., 2011; Rist and Thum, 2017), while Gr28a is found in a largely, but not entirely, distinct set of GRNs. Whether and what kind of Gr genes might be co-expressed with Gr28 in Gr28only GRNs will require more in-depth expression studies and might shed light on other receptors involved in appetitive behaviors of larvae.

A key finding of the work presented here is the observation that Gr28aonlyand Gr28b.c neurons dictate distinct behavioral programs, the former representing an ensemble of neurons that instruct larvae to “go towards” a chemical source and consume it, while the latter do the opposite (Figure 2). This observation is reminiscent of a seminal study by Troemel and colleagues in the C. elegans chemosensory system, who reported that the valence of a chemical compound is dependent on the identity of a neuron, and not the identity of the molecular receptor the neuron expresses (Troemel et al., 1997). The number of “go-away” GRNs in Drosophila larvae co-expressing Gr28b.c and Gr66a is remarkably small, consisting of only two pairs, one in the TO and the other in the DPS/VPS. It seems likely that this is the smallest, minimal subset of neurons sufficient to induce avoidance behavior, as expression of V1R in only the TO pair (under the control of either Gr28b.a-GAL4 or Gr28b.e-GAL4) has no behavioral effect when challenged with capsaicin. The “go-to” neurons are characterized by expression of Gr28a and represent a slightly larger set of four GRN pairs (Gr28aonly GRNs). Thus, the minimal requirement to induce “go-to” and “go-away” behavior is defined by distinct sets of GRNs, and each appears to be composed of neurons located in both external as well as the internal taste organs (Figure 7). Co-expression of the RNA receptor Gr28a and in the DPS/VPS GRN essential for bitter taste (Figures 2B and 3) raises interesting questions about additional functions for Gr28a in bitter taste. We note that the sweet taste receptor Gr5a, a subunit of a multimeric trehalose receptor, is also expressed in non-sweet neurons of unknown function (Fujii et al., 2015).

Functional redundancy of taste receptors

Both behavioral analyses as well as Ca2+ imaging experiments implicate at least one Gr28b protein as an essential component of a denatonium receptor complex, as ΔGr28 larvae exhibit total loss of avoidance (Figure 4), and respective GRNs fail to elicit a response upon exposure to this chemical (Figure 3). What the precise composition of that complex is remains to be determined, but recovery of denatonium responses by expressing either Gr28b.c and/or Gr28b.a indicates that either one of these (or possibly both) is an essential subunit, in addition to other Grs expressed in this GRNs, such as Gr22a and Gr59c (Choi et al., 2020; Rist and Thum, 2017), while Gr66a is unlikely to be part of such a complex based on our behavioral analysis (Figure 4-figure supplement 1).

Since only Gr28b.a and Gr28b.c can rescue denatonium responses in Gr28b.c GRNs of ΔGr28 mutant larvae, sequence comparison between the unique N-terminal halves of the Gr28 proteins comprising the first four transmembrane domains and the extracellular loops 1 and 2 might provide insights as to possible residues important for ligand recognition. When interrogating these regions, only seven residues are identical between Gr28b.a and Gr28b.c (Table 1, Figure 6). Reducing the stringency requirement by allowing one of the remaining receptors to share the same residue, nine additional sites are identified. One experimental avenue to validate these residues as important sites contributing to denatonium binding might involve introduction of point mutations that converts respective amino acids of other Gr28b proteins into those found in Gr28b.a/Gr28b.c.

Amino acid alignment of the six Gr28 proteins.

The sequence alignment was generated using Clustal Omega tool from ClustalW2 (https://www.ebi.ac.uk/Tools/msa/clustalo/). IL and EL indicate intracellular loop and extracellular loop, respectively. Note that the C terminal region starting at the IL3 is identical in the Gr28b proteins. Red highlighted letters indicate amino acids identical only in Gr28b.a and Gr28b.c. Green highlighted letters indicate amino acids conserved in Gr28b.a, Gr28b.c and one other Gr28 protein. Asterisks below the sequences indicate residues identical in all Gr28 proteins, colons indicate conserved residue (STA, NEQK, NHQK, NDEQ, QHRK, MILV, MILF, HY, FYW), and periods indicate moderately conserved residue (CSA, ATV, SAG, STNK, STPA, SGND, SNDEQK, NDEQHK, NEQHRK, FVLIM, HFY). TM1-7 indicate helical transmembrane segments predicted by HMMTOP 2.0 software.

Role of different GRN subsets in taste behavior of larvae

GRNs sufficient for mediating avoidance behavior can be defined by Gr28b.c-GAL4, while GRNs sufficient for mediating appetitive behavior are defined by a subset of Gr28a-GAL4 GRNs (Gr28aonly GRNs; see also Figure 2B and Figure 2-figure supplement 1). Note that each ensemble is composed of at least a pair of neurons located in the external taste organs and a pair of neurons in the internal taste organs. Also, a larger set of avoidance neurons (Gr66a-GAL4) might function independently of any Gr28b.c neurons, and one set of fructose sensing neurons (Gr43aGAL4) distinct from Gr28a-GAL4 GRNs mediates appetitive behavior.

The role of Gr28b proteins in quinine detection is less clear, and the different phenotypes observed in behavioral experiments and Ca2+ imaging suggest that at least two molecular types of quinine receptors exist in larvae. Ca2+ imaging experiments implicate a role for multiple Gr28b subunits in a quinine receptor complex in the TOG neuron, since single Gr28b genes cannot restore the loss of quinine response in ΔGr28 larvae (Figure 5E and 5G). However, because ΔGr28 mutant larvae still avoid quinine (Figure 4), at least one Gr28b independent receptor must be exist in one or several other GRNs, one of which is likely the DPS/VPS, since expression of Kir2.1 in that neuron, along with the TOG neuron, totally abolishes quinine avoidance (Figure 3). We note that functional redundancy is also observed in sweet taste receptors, where we found that different combinations of sugar Gr genes can restore responses to the same sugar when expressed in tarsal sweet GRNs of sugar blind flies (Fujii et al. unpublished).

Materials and methods

Key resources table

Drosophila Stocks

Flies were maintained on standard corn meal food in plastic vials under a 12 h light/dark cycle at 25°C. The w1118 strain (Bloomington Drosophila Stock Center, number 3605) was used as a wild-type control. Fly strains used: Gr28a-GAL4(SF36S) for Figure 1 and Figure 2B, and SF36E1 for Figure 2B and Figure 2-figure supplement 1, Gr28b.a-GAL4(NT42aC51a), Gr28b.c-GAL4(NT21B1) (Thorne and Amrein, 2008); Gr28b.e-GAL4(Gr28a3AII) and Gr66a-GAL4 (Scott et al., 2001); ΔGr28(54B3) (Mishra et al., 2018); UAS-Gr28a, UAS-Gr28b.a, UAS-Gr28b.b, UAS-Gr28b.c, UAS-Gr28b.d and UAS-Gr28b.e (Ni et al., 2013); Gr43aGAL4 (Miyamoto et al., 2012); Gr66a-LexA and lexAop-GAL80 (Thistle et al., 2012); UAS-VR1E600K (Marella et al., 2006); UAS-Kir2.1-GFP (Baines et al., 2001; Paradis et al., 2001); lexAop-rCD2:GFP (Lai and Lee, 2006), UAS-GCaMP6m, UAS-mCD8:RFP and Gr66aex83 (Bloomington Drosophila Stock Center, numbers 42748, 32220 and 35528); Gr28b.c-LexA(#8)

Chemicals

Caffeine (Cat No. C0750), capsaicin (Cat No. M2028), denatonium benzoate (Cat No. D5765) and lobeline hydrochloride (Cat No. 141879), D-(-)-ribose (Cat No. R7500) and quinine hydrochloride dihydrate (Cat No. Q1125) were purchased from Millipore-Sigma, with a purity of > 95%. Fructose (Cat No. F1092) and agarose (Cat. No. 20-102) were purchased from Spectrum chemical and Apexbio, respectively. Sucrose (Mfr. No. 8360-06) and charcoal (Cat No. 1560-01) were purchased from Macron fine chemicals and J.T. Baker, respectively. A stock solution for capsaicin (20 mM) was prepared in 70% ethanol and stored at 4° C protected from light for up to one year. Stock solutions for bitter chemicals were prepared in Millipore Q water and stored at -20° C. Stock solutions for sugars were prepared in Millipore Q water and stored at 4° C for up to one month. A stock solution for ribose was treated with charcoal (10% of the weight of ribose used for stock solution) overnight at 4° C and sterile-filtrated (0.45 μm) to remove unrelated odor. Stock solutions were diluted to the final concentration using Millipore Q water prior to each experiment.

Immunofluorescence

Immunofluorescence of larval heads was performed based on the protocol described in Croset and colleagues (Croset et al., 2016) with minor modification. Heads of third instar larvae were dissected using microscissors in phosphate buffered saline (PBS) and immediately fixed in PBS with 4% paraformaldehyde for 1 hour at 4 °C. They were washed six times in washing buffer (PBS with 0.1% Triton X-100) for 20 min and blocked for 1 h in washing buffer containing 5% heat-inactivated goat serum (SouthernBiotech, Cat. No. 0060-01), followed by incubation with the primary antibodies (rabbit anti–GFP, 1:1,000 dilution; rat anti–mCD8, 1:200 dilution, ThermoFisher Scientific) at 4°C overnight. The next day, heads were washed six times for 20 min in washing buffer and blocked in washing buffer containing 5% heat-inactivated goat serum for 1 hour, followed by incubation with the secondary antibodies (goat anti rabbit ALEXA 488, 1:500 dilution, ThermoFisher Scientific; goat anti–rat Cy3, 1:300 dilution, Jackson Immunoresearch Laboratories Inc.) at 4°C overnight. Finally, heads were washed six times in washing buffer for 20 min each at room temperature under gentle agitation. Heads were then mounted with VectaShield (Vector Lab, Cat No. H-1200) on a microscope slide and images were obtained using a Nikon A1R confocal microscope system. Adobe photoshop 2022 was used further to process images.

Larval two-choice preference assay

Two-choice preference assay of larvae was conducted as described in Mishra et al. (Mishra et al., 2013) with minor modifications. Flies were placed on standard corn meal food in plastic vials and allowed to lay eggs for 24 h under a 12 h light/dark cycle at 25°C. Flies were removed from food vials and feeding-stage third-instar larvae were collected. Agarose food dishes for two-choice preference assay were prepared just prior each experiment as follows: Petri dishes (60 x 15 mm, Falcon, Cat. No. REF353004) with two halves marked on the bottom were filled with melted plain 1% agarose or 1% agarose containing 1.75% ethanol (for capsaicin preference). After the agarose solidified, one half was removed and replaced with 1% agarose solution containing taste ligands (capsaicin or bitter compound). For each experiment, 15 larvae from food vials were briefly rinsed twice with Millipore Q water and placed along the middle separating pure and ligand containing agarose. After 16 min, images were taken for record keeping and used to calculate larval preference indices. Larvae that crawled onto the wall of a dish or dug in the agarose were excluded. The preference index (P.I.) was calculated as follow: PI = (Ntastant - Nplain) / NTotal, whereby N is the number of larvae in the tastant sector, the plain agarose sector and the total number, respectively. Positive values indicate a preference for capsaicin or bitter compound while negative values indicate repulsion (avoidance).

Calcium imaging

Calcium imaging was performed in Gr28b.c GRNs expressed in the terminal organ of feeding-stage, third instar larvae, reared as described for the larval two-choice preference assay. For each experiment, larvae from food vials were briefly rinsed twice with Millipore Q water and were mounted dorsally on a large microscope cover glass (24 x 50 mm, VWR, Cat. No. 16004-098) using double-sided scotch tape and covered with a small microscope cover glass (12CIR-1, Thermo Fisher Scientific, Cat. No. 1254580). Millipore Q water (40 µl) was applied to the tip of the larval head, and the preparation was placed on the stage of a Nikon eclipse Ti inverted microscope. Images were obtained every 500 ms, starting 15 s before application and ending 105 s after ligand application. Each recording was initiated by applying water (40μl) to set a baseline. The first ligand solution (40 μl of bitter chemical or sugar) was applied thereafter, followed by 5 washes with carrier (100 μl of water). After a 3-minute pause to allow the preparation to recalibrate, a second ligand solution (40 μl bitter chemical or sugar) was applied. To assure validity in experiments with Gr28 mutants and rescues, each recording was concluded with application of caffeine, and recordings were included only if caffeine generated a positive response. Baseline fluorescence, which was determined from the average of five frame measurements from a region next to the cell immediately before ligand application, was subtracted from the actual measurements. ΔF/F (%) = (fluorescence light intensity of the cell body – baseline/baseline) x 100. ΔF/F (max %) is the maximum value within 40 seconds after ligand application.

Generation of transgenic Gr28b.c-LexA flies

To generate the Gr28b.c-LexA driver, a 1.3 kb DNA fragment immediately upstream of the Gr28b.c start codon was amplified from w1118 flies using a forward (5′-AATCTAGGTACCCCGGCTGCTCGTCTCCCTGGATGT-3′) and a reverse (5′- CGTCAAACTAGTGACCGCTTCGTTTGAGCTTCAACC-3′) primer. Acc65I and SpeI sites included in the primer sequence (underlined) were incorporated such that the amplified fragment was amenable to directional cloning into the LexA vector CMC105 (Larsson et al., 2004). The clone chosen was confirmed by DNA sequence analysis. Transgenic flies were generated by standard P-element transformation of w1118 embryos (Rainbow Transgenic Flies Inc., Camarillo, CA).

Statistical analysis

Statistical analyses were conducted using Prism software 9.5.1 (GraphPad Software). Larval two-choice preference assay and Ca2+ imaging data were analyzed for normal distribution using D’Agostino-Pearson omnibus and Shapiro-Wilk normality tests. When groups did not meet the assumption for normal distribution, nonparametric statistics was used. For comparison between multiple groups, one-way ANOVA or Kruskal-Wallis test by ranks (nonparametric one-way ANOVA) was performed to test for difference of mean or rank distribution. As a post hoc test, Bonferroni’s or Dunn’s (nonparametric) multiple comparison tests were employed to compare two specific groups. One-way ANOVA with Bonferroni’s multiple comparison tests were used in Figure 2. Kruskal-Wallis test by ranks with Dunn’s multiple comparison tests were used in Figures 3, 4B and 5F and 5G). For comparison between two groups, Mann-Whitney U test (nonparametric t test, Figures 4A, 5D, 5E, Figure 2-figure supplement 1B and Figure 4-figure supplement 1) with two-tailed P value were used to compare means of two groups. Sample size for larval two-choice preference assays and Ca2+ imaging experiments were based on Mishra et al (Mishra et al., 2018).

Supporting information